---
title: "Age-Related Changes in Tissue Iron and Relationships with Substance Use"
author: Ashley C. Parr
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: pdf_document
---

# TISSUE IRON  

```{r, include=FALSE}

if(!is.null(dev.list())) dev.off()# Clear plots
cat("\014") # Clear console
rm(list=ls())# Clean workspace

library(rmarkdown)
library(tinytex)
library(dplyr)
library(tidyr)
library(ggplot2)
library(LNCDR)
library(mgcv)
library(readr)
library(RColorBrewer)
library(ggpubr)
library(plotrix)
library(ggeffects)

knitr::opts_chunk$set(out.width="50%", out.height="50%", warning = FALSE, 
                      fig.pos = 'H', dev = 'png', dpi=150)


library(showtext)
font_add("Arial", "/System/Library/Fonts/Supplemental/Arial.ttf")  # Use the actual file path
showtext_auto()

```      

# Housekeeping
## Bring in Relevant Files, Merge, & Clean  
```{r, echo=FALSE}

All_Sites_Iron <- "/Users/ashparr/Library/Mobile Documents/com~apple~CloudDocs/Documents/Pitt/Manuscripts & Projects/NCANDA Project/DATA/ALL_SITES_IRON_CLEANED_y8.csv"
All_Sites_Iron <- read.csv(All_Sites_Iron)

All_Sites_Substance <- "/Users/ashparr/Library/Mobile Documents/com~apple~CloudDocs/Documents/Pitt/Manuscripts & Projects/NCANDA Project/DATA/ALL_SITES_SUBSTANCE_CLEANED_y8_newnicotine_08042025.RData"
All_Sites_Substance1 <- load(All_Sites_Substance)

Long_Substance <- "/Users/ashparr/Library/Mobile Documents/com~apple~CloudDocs/Documents/Pitt/Manuscripts & Projects/NCANDA Project/DATA/LONG_SUBSTANCE_y8_newnicotine_08042025.RData"
Long_Substance1 <- load(Long_Substance)

# # #Get rid of 'all' since using probabilistic score 
Long_Substance <- Long_Substance %>% filter(Substance!="All" & Substance !="All Mean")

Long_Substance$Substance <- droplevels(Long_Substance$Substance)
Long_Substance$oSubstance <- droplevels(Long_Substance$oSubstance)


# Clean Iron 

#RM stuff in all sites substance
All_Sites_Iron <- dplyr::select(All_Sites_Iron, subject, visit, roi, site, sex, visit_age, 
                                tat2.combat, tat2.combat_z)

All_Sites_Iron$roi[All_Sites_Iron$roi=="striatum"] <- "BG"
All_Sites_Iron$roi[All_Sites_Iron$roi=="pallidum"] <- "Pallidum"
All_Sites_Iron$roi[All_Sites_Iron$roi=="caudate"] <- "Caudate"
All_Sites_Iron$roi[All_Sites_Iron$roi=="nacc"] <- "NAcc"
All_Sites_Iron$roi[All_Sites_Iron$roi=="putamen"] <- "Putamen"

columns_to_factorize <- colnames(All_Sites_Iron)[grep("high", colnames(All_Sites_Iron))]
 
# Factorize columns
for (col in columns_to_factorize) {
  All_Sites_Iron[[col]] <- factor(All_Sites_Iron[[col]])
} 
 
All_Sites_Iron$site <- as.factor(All_Sites_Iron$site)
All_Sites_Iron$subject <- as.factor(All_Sites_Iron$subject)
All_Sites_Iron$sex<- as.factor(All_Sites_Iron$sex)
All_Sites_Iron$oROI <- ordered(All_Sites_Iron$roi, levels = c("BG", "NAcc","Putamen",
                                                              "Caudate","Pallidum"))
 

All_Sites_Iron$oROI2 <- ordered(All_Sites_Iron$oROI, levels = c( "BG","NAcc", "Pallidum","Putamen", 
                                                                          "Caudate"))

All_Sites_Iron_Long <- merge(All_Sites_Iron, Long_Substance, by = c("subject", "visit",
                                                                    "site", "sex",
                                                                    "visit_age"),
                             all.y = TRUE)%>% filter(!is.na(tat2.combat))


All_Sites_Iron <- merge(All_Sites_Iron, All_Sites_Substance, by = c("subject","visit",
                                                                    "site", "sex", 
                                                                    "visit_age"),
                        all.y = TRUE) %>% filter(!is.na(tat2.combat))

  
```    
 
## Calculate Summary Stats: Mean, Peak ages, Peak nT2*w, ETC. Across full sample & Demographic Factors 

```{r, echo=FALSE, fig.align = "center"}

# Note: peak is MIN for iron since inverse relationship between nT2*w and iron content 

All_Sites_Iron <- All_Sites_Iron %>%
  group_by(subject, oROI) %>%
    dplyr::mutate(
    valid_scores = !is.na(tat2.combat),
    peak_age_Iron = if(any(valid_scores)) visit_age[which.min(ifelse(valid_scores, tat2.combat, -Inf))] else NA_real_,
    peak_Iron = if(any(valid_scores)) min(tat2.combat, na.rm = TRUE) else NA_real_  # Changed this line!
  ) %>% 
  dplyr::select(-valid_scores)


subscale_summary <- All_Sites_Iron %>%
  group_by(oROI) %>%
  reframe(
    # Mean and SD of all scores for this subscale
    mean_Iron = mean(tat2.combat[oROI == cur_group()$oROI], na.rm = TRUE),
    sd_Iron = std.error(tat2.combat[oROI == cur_group()$oROI], na.rm = TRUE),
    # Mean of peak scores across subjects
    mean_peak_Iron = mean(peak_Iron, na.rm = TRUE),
    sd_peak_Iron = std.error(peak_Iron, na.rm = TRUE),
    # Mean age at which peak occurs
    mean_peak_age_Iron = mean(peak_age_Iron, na.rm = TRUE),
    SD_peak_age_Iron = std.error(peak_age_Iron, na.rm = TRUE),
    # Additional useful stats
    n_subjects = n_distinct(subject),
    n_valid_peaks = sum(!is.na(peak_Iron)),
    .groups = "drop"
  )
  

# subscale_summary[] <- lapply(subscale_summary, function(x) if (is.numeric(x)) round(x, 2) else x)

write.csv(subscale_summary, file='/Users/ashparr/Library/Mobile Documents/com~apple~CloudDocs/Documents/Pitt/Manuscripts & Projects/NCANDA Project/DATA/Summary_Table_Across_IronROIs_FullSample.csv', row.names = FALSE)


subscale_summary_sex <- All_Sites_Iron %>%
  group_by(oROI, sex) %>%
  reframe(
    # Mean and SD of all scores for this subscale
    mean_Iron = mean(tat2.combat[oROI == cur_group()$oROI], na.rm = TRUE),
    sd_Iron = std.error(tat2.combat[oROI == cur_group()$oROI], na.rm = TRUE),
    # Mean of peak scores across subjects
    mean_peak_Iron = mean(peak_Iron, na.rm = TRUE),
    sd_peak_Iron = std.error(peak_Iron, na.rm = TRUE),
    # Mean age at which peak occurs
    mean_peak_age_Iron = mean(peak_age_Iron, na.rm = TRUE),
    SD_peak_age_Iron = std.error(peak_age_Iron, na.rm = TRUE),
    # Additional useful stats
    n_subjects = n_distinct(subject),
    n_valid_peaks = sum(!is.na(peak_Iron)),
    .groups = "drop"
  )
  

subscale_summary_income <- All_Sites_Iron %>%
  group_by(oROI, oIncome_recode) %>%
  reframe(
    # Mean and SD of all scores for this subscale
    mean_Iron = mean(tat2.combat[oROI == cur_group()$oROI], na.rm = TRUE),
    sd_Iron = std.error(tat2.combat[oROI == cur_group()$oROI], na.rm = TRUE),
    # Mean of peak scores across subjects
    mean_peak_Iron = mean(peak_Iron, na.rm = TRUE),
    sd_peak_Iron = std.error(peak_Iron, na.rm = TRUE),
    # Mean age at which peak occurs
    mean_peak_age_Iron = mean(peak_age_Iron, na.rm = TRUE),
    SD_peak_age_Iron = std.error(peak_age_Iron, na.rm = TRUE),
    # Additional useful stats
    n_subjects = n_distinct(subject),
    n_valid_peaks = sum(!is.na(peak_Iron)),
    .groups = "drop"
  )


subscale_summary_race <- All_Sites_Iron %>%
  group_by(oROI, oRaceeth_recode) %>%
  reframe(
    # Mean and SD of all scores for this subscale
    mean_Iron = mean(tat2.combat[oROI == cur_group()$oROI], na.rm = TRUE),
    sd_Iron = std.error(tat2.combat[oROI == cur_group()$oROI], na.rm = TRUE),
    # Mean of peak scores across subjects
    mean_peak_Iron = mean(peak_Iron, na.rm = TRUE),
    sd_peak_Iron = std.error(peak_Iron, na.rm = TRUE),
    # Mean age at which peak occurs
    mean_peak_age_Iron = mean(peak_age_Iron, na.rm = TRUE),
    SD_peak_age_Iron = std.error(peak_age_Iron, na.rm = TRUE),
    # Additional useful stats
    n_subjects = n_distinct(subject),
    n_valid_peaks = sum(!is.na(peak_Iron)),
    .groups = "drop"
  )



subscale_summary_site <- All_Sites_Iron %>%
  group_by(oROI, site) %>%
  reframe(
    # Mean and SD of all scores for this subscale
    mean_Iron = mean(tat2.combat[oROI == cur_group()$oROI], na.rm = TRUE),
    sd_Iron = std.error(tat2.combat[oROI == cur_group()$oROI], na.rm = TRUE),
    # Mean of peak scores across subjects
    mean_peak_Iron = mean(peak_Iron, na.rm = TRUE),
    sd_peak_Iron = std.error(peak_Iron, na.rm = TRUE),
    # Mean age at which peak occurs
    mean_peak_age_Iron = mean(peak_age_Iron, na.rm = TRUE),
    SD_peak_age_Iron = std.error(peak_age_Iron, na.rm = TRUE),
    # Additional useful stats
    n_subjects = n_distinct(subject),
    n_valid_peaks = sum(!is.na(peak_Iron)),
    .groups = "drop"
  )


subscale_summary_sex <- subscale_summary_sex %>%
  dplyr::rename(Label = sex)
subscale_summary_income <- subscale_summary_income %>%
  dplyr::rename(Label = oIncome_recode)
subscale_summary_race <- subscale_summary_race %>%
  dplyr::rename(Label = oRaceeth_recode)
subscale_summary_site <- subscale_summary_site %>%
 dplyr::rename(Label = site)
subscale_summary$Label <- "All"

Long_means_all_SES <- rbind(subscale_summary_sex, subscale_summary_income, 
                            subscale_summary_race, subscale_summary_site,subscale_summary)



# Long_means_all_SES[] <- lapply(Long_means_all_SES, function(x) if (is.numeric(x)) round(x, 2) else x)


custom_order <- c("BG", "Caudate", "NAcc", "Pallidum", "Putamen")

# Convert to factor with levels in custom order
Long_means_all_SES <- Long_means_all_SES %>%
  mutate(oROI = factor(oROI, levels = custom_order)) %>%
  arrange(oROI)
 

write.csv(Long_means_all_SES, file='/Users/ashparr/Library/Mobile Documents/com~apple~CloudDocs/Documents/Pitt/Manuscripts & Projects/NCANDA Project/DATA/Summary_Table_Across_IronROIs_SES.csv', row.names = FALSE)
 
Nsubs <- dplyr::select(All_Sites_Iron[!is.na(All_Sites_Iron$tat2.combat),], visit, visit_age, subject, sex)
Nsubs <- unique(Nsubs)


``` 

## Clean Environment 
```{r, echo = FALSE, fig.align='center'}

rm(list=ls()[! ls() %in% c("All_Sites_Iron","All_Sites_Substance","Five_Cat_SU_Iron",
                           "Three_Cat_SU_Iron", "Five_Cat_SU", "Three_Cat_SU",
                           "Anti_Saccade","Anti_Saccade_Iron", "All_Sites_Iron_Long", 
                           "demolistdf","demolist","NCANDAwithalldemfortable",
                           "SUlistdf", "SUlist", "NCANDAwithallSUfortable", 
                           "nT2listdf", "nT2list", "NCANDAwithallnT2fortable","Long_Substance",
                           "peakdens.first_age","peakdens.reg_age",
                           "NAcc_Iron_Substance_Mod", "Alcohol_Iron_Substance_Mod", "Binge_Iron_Substance_Mod","Long_Iron_means","Long_Iron_means_sex","Long_Iron_means_income","Long_Iron_means_race", 
                           "Long_Iron_means_site", "Long_Iron_grmeans", "peakdens.peak_age", "Long_Iron_peak_means_sex",
                           "Long_Iron_peak_means_income","Long_Iron_peak_means_race","Long_Iron_peak_means_site",
                           "Long_peak_means_all_SES", "Long_means_all_SES", "Demo_All_violin", "subscale_summary"
                           )])


``` 

## Set colors  
```{r, echo = FALSE, fig.align='center'}

substance_colorbar <- c("#260C51", "#EA464C", "#7B02A8", "#079287", "#EAB720")
substance_colorbar2 <- c("#260C51", "#EA464C", "#079287", "#EAB720", "#7B02A8")


iron_colorbar <- c("#461248","#A02F7B","#7F4598","#C196C4","#c47c9e")
iron_colorbar2 <- c("#461248","#A02F7B","#c47c9e","#7F4598","#C196C4")



library(scales)
Demo_Colors <- colorRampPalette(c("#F5F5F9","#461248"))(length(unique(Long_means_all_SES$Label)))

``` 

## Subset DFs
```{r, echo = FALSE, fig.align='center'}

set.seed(1) #for consistency in derivatives + derivative credible intervals

 
### Subset DF 
Iron <- All_Sites_Iron %>% dplyr::select(subject, visit, sex, visit_age, site,
                                                    oIncome_recode,oRaceeth_recode,
                                                    tat2.combat, roi, oROI,oROI2) %>% filter (!is.na(oROI))
Iron$roi <- as.factor(Iron$roi)


``` 

# Test for Differences in nT2*w Magnitude/Mean/Peak across ROIs and Demographic Factors 
## Full Sample
```{r, echo = FALSE, fig.align='center'}

# Across Substances 
gam.model_Iron <- mgcv::gamm(
  tat2.combat ~ s(visit_age, k = 4, fx = F, bs = "cs") + oROI + sex + oIncome_recode + oRaceeth_recode + site,
  random = list(subject = ~1, subject = ~0 + visit_age),  # <- combined intercept & slope
  data = Iron %>% 
    filter(oROI != "BG"),
  method = "REML")
summary(gam.model_Iron$gam)
anova(gam.model_Iron$gam)  
 
``` 

### Plots: 
```{r, echo = FALSE, fig.align='center'}
 
# Violin plot showing distribution + Mean 
Iron_violin <- All_Sites_Iron %>%
  ggplot( aes(x=oROI2, y=tat2.combat, fill=oROI2, color=oROI2)) +
  scale_color_manual(values=iron_colorbar2)+ 
   scale_fill_manual(values=iron_colorbar2)+  
    geom_violin(width=1, size=0.4,adjust = 2) +
      stat_summary(fun = mean, geom = "point", shape = 15, size = 3, color = "black") +   
    theme(aspect.ratio=1.5, legend.position = "none",
         axis.text = element_text(size = 8, family = "Arial", color = c("black")),
  axis.title.x = element_text(size = 9, family ="Arial", color = c("black")),
  axis.title.y = element_text(size = 9, family ="Arial", color = c("black")),
  axis.line = element_line(linewidth = .2), 
  axis.ticks = element_line(linewidth = .2),
        panel.grid.major = element_blank(),panel.grid.minor=element_blank(),
        panel.background=element_blank()) +
  labs(y = "\nnT2*w",
       x = "ROI\n") +
  scale_x_discrete(
    breaks = c("BG", "NAcc", "Pallidum", "Putamen", "Caudate"),
    labels = c("BG", "NAcc", "GP","Put.","Caud."))+
    coord_flip() + # This switch X and Y axis and allows to get the horizontal version
    scale_y_continuous(trans = "reverse")
print(Iron_violin)

peakdens.peak_age = dplyr::select(All_Sites_Iron[!is.na(All_Sites_Iron$peak_age_Iron),],
                             oROI, peak_age_Iron) %>%
                               group_by(oROI) %>%
                               summarise(Max_Dense = density(peak_age_Iron)$x[which.max(density(peak_age_Iron, na.rm=T)$y)])

# Density plot showing distributions of peak ages 
Iron_dense_peak <- ggplot(All_Sites_Iron, aes(x = peak_age_Iron, group = oROI, colour = oROI)) +
  scale_color_manual(values=iron_colorbar)+
  geom_density(linewidth = 1, bw = .80) + 
   theme(aspect.ratio=1.5, legend.position = "none",
         axis.text = element_text(size = 8, family = "Arial", color = c("black")),
  axis.title.x = element_text(size = 9, family ="Arial", color = c("black")),
  axis.title.y = element_text(size = 9, family ="Arial", color = c("black")),
  axis.line = element_line(linewidth = .2), 
  axis.ticks = element_line(linewidth = .2),
        panel.grid.major = element_blank(),panel.grid.minor=element_blank(),
        panel.background=element_blank()) +
    scale_x_continuous(limits = c(10, 30), breaks = seq(10, 30, by = 10))+
    # geom_vline(data = peakdens.peak_age, aes(xintercept=Max_Dense,color = oROI), linetype="dashed") +
         labs(x = "\nAge of Peak nT2*w",
       y = "Density\n")
plot(Iron_dense_peak)
 
Iron_Dists <- ggarrange(Iron_violin, Iron_dense_peak, 
                    ncol = 3, nrow = 1)
Iron_Dists

ggsave(Iron_Dists, filename = "/Users/ashparr/Documents/Pitt/Manuscripts & Projects/NCANDA Project/Results 2025/Iron_Distributions_AllROIs.pdf", device = "pdf", dpi = 500,
       width = 8, height = 8, units = "in")
    
subscale_summary <- subscale_summary[!is.na(subscale_summary$oROI),]
subscale_summary$oROI<-ordered(subscale_summary$oROI,
levels=c("BG", "NAcc", "Putamen","Pallidum", "Caudate"))

  
#Mean Iron
p1_5 <- subscale_summary %>% ungroup () %>%  
  ggplot(aes(x = mean_Iron, y = oROI, fill = oROI)) +  
  geom_point(shape = 21, size = 5.5, color = "white") +
        scale_x_continuous(trans = "reverse", limits = c(.015, .010), breaks = c(.015, .0125, .010))+ 
  scale_fill_manual(values = iron_colorbar2) +
  labs(x = "\nnT2*w",
       y = "ROI\n") +
  scale_y_discrete(
    breaks = c("BG", "NAcc", "Pallidum", "Putamen", "Caudate"),
    labels = c("BG", "NAcc", "GP","Put.","Caud."))+
  theme(aspect.ratio = 1.5,
  legend.position = "none",
  axis.text = element_text(size = 8, family = "Arial", color = c("black")),
  axis.title.x = element_text(size = 9, family ="Arial", color = c("black")),
  axis.title.y = element_text(size = 9, family ="Arial", color = c("black")),
  axis.line = element_line(linewidth = .2), 
  axis.ticks = element_line(linewidth = .2),
  panel.grid.major = element_blank(),
  panel.grid.minor=element_blank(),
  panel.background=element_blank())
print(p1_5)
 
#Mean Peak Iron
p2_5 <- subscale_summary %>% ungroup () %>%  
  ggplot(aes(x = mean_peak_Iron, y = oROI, fill = oROI)) +  
  geom_point(shape = 21, size = 5.5, color = "white") +
          scale_x_continuous(trans = "reverse", limits = c(.015, .009), breaks = c(.015, .0125, .010))+ 
  scale_fill_manual(values = iron_colorbar2) +
  labs(x = "\nPeak nT2*w",
       y = "ROI\n") +
  scale_y_discrete(
    breaks = c("BG", "NAcc", "Pallidum", "Putamen", "Caudate"),
    labels = c("BG", "NAcc", "GP","Put.","Caud."))+
  theme(aspect.ratio = 1.5,
  legend.position = "none",
  axis.text = element_text(size = 8, family = "Arial", color = c("black")),
  axis.title.x = element_text(size = 9, family ="Arial", color = c("black")),
  axis.title.y = element_text(size = 9, family ="Arial", color = c("black")),
  axis.line = element_line(linewidth = .2), 
  axis.ticks = element_line(linewidth = .2),
  panel.grid.major = element_blank(),
  panel.grid.minor=element_blank(),
  panel.background=element_blank())
print(p2_5)


#Mean Age of Peak Iron
p3_5 <- subscale_summary %>% ungroup () %>%  
  ggplot(aes(x = mean_peak_age_Iron, y = oROI, fill = oROI)) +  
  geom_point(shape = 21, size = 5.5, color = "white") +
          scale_x_continuous(limits = c(18, 22), breaks = seq(18, 22, by = 2))+
  scale_fill_manual(values = iron_colorbar2) +
  labs(x = "\nAge of Peak nT2*w",
       y = "ROI\n") +
  scale_y_discrete(
    breaks = c("BG", "NAcc", "Pallidum", "Putamen", "Caudate"),
    labels = c("BG", "NAcc", "GP","Put.","Caud."))+
  theme(aspect.ratio = 1.5,
  legend.position = "none",
  axis.text = element_text(size = 8, family = "Arial", color = c("black")),
  axis.title.x = element_text(size = 9, family ="Arial", color = c("black")),
  axis.title.y = element_text(size = 9, family ="Arial", color = c("black")),
  axis.line = element_line(linewidth = .2), 
  axis.ticks = element_line(linewidth = .2),
  panel.grid.major = element_blank(),
  panel.grid.minor=element_blank(),
  panel.background=element_blank())
print(p3_5)


Iron_Summ <- ggarrange(p1_5, p2_5, p3_5, 
                    ncol = 3, nrow = 1)
Iron_Summ

ggsave(Iron_Summ, filename = "/Users/ashparr/Documents/Pitt/Manuscripts & Projects/NCANDA Project/Results 2025/Iron_MeanPeakPAge_AllROIs.pdf", device = "pdf", dpi = 500,
       width = 8, height = 8, units = "in")


#Mean & SE Iron
p1_5_SE <- ggplot(subscale_summary, aes(mean_Iron, oROI)) +
  geom_pointrange(
    aes(xmin = mean_Iron-sd_Iron, xmax = mean_Iron+sd_Iron, color = oROI), position = position_dodge(0.3))+ 
  scale_x_continuous(trans = "reverse", limits = c(.015, .010), breaks = c(.015, .0125, .010))+ 
  scale_color_manual(values = iron_colorbar2) +
  scale_fill_manual(values = iron_colorbar2) +
  labs(x = "\nnT2*w",
       y = "ROI\n") +
  scale_y_discrete(
    breaks = c("BG", "NAcc", "Pallidum", "Putamen", "Caudate"),
    labels = c("BG", "NAcc", "GP","Put.","Caud."))+
  theme(aspect.ratio = 1.5,
  legend.position = "none",
  axis.text = element_text(size = 8, family = "Arial", color = c("black")),
  axis.title.x = element_text(size = 9, family ="Arial", color = c("black")),
  axis.title.y = element_text(size = 9, family ="Arial", color = c("black")),
  axis.line = element_line(linewidth = .2), 
  axis.ticks = element_line(linewidth = .2),
  panel.grid.major = element_blank(),
  panel.grid.minor=element_blank(),
  panel.background=element_blank())
print(p1_5_SE)
 
  
#Mean & SE Peak Iron
p2_5_SE <- ggplot(subscale_summary, aes(mean_peak_Iron, oROI)) +
  geom_pointrange(
    aes(xmin = mean_peak_Iron-sd_peak_Iron, xmax = mean_peak_Iron+sd_peak_Iron, color = oROI), position = position_dodge(0.3))+ 
scale_x_continuous(trans = "reverse", limits = c(.015, .009), breaks = c(.015, .0125, .010))+
  scale_color_manual(values = iron_colorbar2) +
  scale_fill_manual(values = iron_colorbar2) +
 labs(x = "\nPeak nT2*w",
       y = "ROI\n") +
  scale_y_discrete(
    breaks = c("BG", "NAcc", "Pallidum", "Putamen", "Caudate"),
    labels = c("BG", "NAcc", "GP","Put.","Caud."))+
  theme(aspect.ratio = 1.5,
  legend.position = "none",
  axis.text = element_text(size = 8, family = "Arial", color = c("black")),
  axis.title.x = element_text(size = 9, family ="Arial", color = c("black")),
  axis.title.y = element_text(size = 9, family ="Arial", color = c("black")),
  axis.line = element_line(linewidth = .2), 
  axis.ticks = element_line(linewidth = .2),
  panel.grid.major = element_blank(),
  panel.grid.minor=element_blank(),
  panel.background=element_blank())
print(p2_5_SE)


#Mean & SE Age of Peak Iron
p3_5_SE <- ggplot(subscale_summary, aes(mean_peak_age_Iron, oROI)) +
  geom_pointrange(
    aes(xmin = mean_peak_age_Iron-SD_peak_age_Iron, xmax = mean_peak_age_Iron+SD_peak_age_Iron, color = oROI), position = position_dodge(0.3))+ 
            scale_x_continuous(limits = c(18, 22), breaks = seq(18, 22, by = 2))+ 
  scale_fill_manual(values = iron_colorbar2) +
  scale_color_manual(values = iron_colorbar2) +
  labs(x = "\nAge of Peak nT2*w",
       y = "ROI\n") +
  scale_y_discrete(
    breaks = c("BG", "NAcc", "Pallidum", "Putamen", "Caudate"),
    labels = c("BG", "NAcc", "GP","Put.","Caud."))+
  theme(aspect.ratio = 1.5,
  legend.position = "none",
  axis.text = element_text(size = 8, family = "Arial", color = c("black")),
  axis.title.x = element_text(size = 9, family ="Arial", color = c("black")),
  axis.title.y = element_text(size = 9, family ="Arial", color = c("black")),
  axis.line = element_line(linewidth = .2), 
  axis.ticks = element_line(linewidth = .2),
  panel.grid.major = element_blank(),
  panel.grid.minor=element_blank(),
  panel.background=element_blank())
print(p3_5_SE)


Iron_Summ_SE <- ggarrange(p1_5_SE, p2_5_SE, p3_5_SE, 
                    ncol = 3, nrow = 1)
Iron_Summ_SE

ggsave(Iron_Summ_SE, filename = "/Users/ashparr/Documents/Pitt/Manuscripts & Projects/NCANDA Project/Results 2025/Iron_MeanPeakPAge_SE_AllROIs.pdf", device = "pdf", dpi = 500,
       width = 8, height = 8, units = "in")

``` 

## Demographic Factors
```{r, echo = FALSE, fig.align='center'}

# Initialize empty results data frame
Iron_Results.sex <- data.frame(
  oROI = character(),
  F_value = numeric(),
  p_value = numeric(),
  stringsAsFactors = FALSE
)

# Loop through each level of oSubstance
for (s in unique(Iron$oROI)) {
  
  # Subset data for the current substance
  sub_df <- Iron %>% filter(oROI == s)
  # Try fitting the model
  model <- mgcv::gamm(
      tat2.combat ~ s(visit_age, k = 4, fx = FALSE, bs = "cs") + sex,
      random = list(subject = ~1, subject = ~0 + visit_age),
      data = sub_df,
      method = "REML")
  
  # Run ANOVA on the GAM component
  aov_result <- anova(model$gam)
    F_val <- aov_result$pTerms.table[2]
    p_val <- aov_result$pTerms.table[3]
    
    # Append to results
    Iron_Results.sex <- rbind(Iron_Results.sex, data.frame(
      oROI = s,
      F_value = round(F_val, 2),
      p_value = signif(p_val, 2)
    ))
  }


Iron_Results.sex$Label <- "Sex"


# Initialize empty results data frame
Iron_Results.income <- data.frame(
  oROI = character(),
  F_value = numeric(),
  p_value = numeric(),
  stringsAsFactors = FALSE
)

# Loop through each level of oSubstance
for (s in unique(Iron$oROI)) {
  
  # Subset data for the current substance
  sub_df <- Iron %>% filter(oROI == s)
  # Try fitting the model
  model <- mgcv::gamm(
      tat2.combat ~ s(visit_age, k = 4, fx = FALSE, bs = "cs") + oIncome_recode,
      random = list(subject = ~1, subject = ~0 + visit_age),
      data = sub_df,
      method = "REML")
  
  # Run ANOVA on the GAM component
  aov_result <- anova(model$gam)
    F_val <- aov_result$pTerms.table[2]
    p_val <- aov_result$pTerms.table[3]
    
    # Append to results
    Iron_Results.income <- rbind(Iron_Results.income, data.frame(
      oROI = s,
      F_value = round(F_val, 2),
      p_value = signif(p_val, 2)
    ))
  }

Iron_Results.income$Label <- "Income"


# Initialize empty results data frame
Iron_Results.raceeth <- data.frame(
  oROI = character(),
  F_value = numeric(),
  p_value = numeric(),
  stringsAsFactors = FALSE
)

# Loop through each level of oSubstance
for (s in unique(Iron$oROI)) {
  
  # Subset data for the current substance
  sub_df <- Iron %>% filter(oROI == s)
  # Try fitting the model
  model <- mgcv::gamm(
      tat2.combat ~ s(visit_age, k = 4, fx = FALSE, bs = "cs") + oRaceeth_recode,
      random = list(subject = ~1, subject = ~0 + visit_age),
      data = sub_df,
      method = "REML")
  
  # Run ANOVA on the GAM component
  aov_result <- anova(model$gam)
    F_val <- aov_result$pTerms.table[2]
    p_val <- aov_result$pTerms.table[3]
    
    # Append to results
    Iron_Results.raceeth <- rbind(Iron_Results.raceeth, data.frame(
      oROI = s,
      F_value = round(F_val, 2),
      p_value = signif(p_val, 2)
    ))
  }

Iron_Results.raceeth$Label <- "Race/Eth"



# Initialize empty results data frame
Iron_Results.site <- data.frame(
  oROI = character(),
  F_value = numeric(),
  p_value = numeric(),
  stringsAsFactors = FALSE
)

# Loop through each level of oSubstance
for (s in unique(Iron$oROI)) {
  
  # Subset data for the current substance
  sub_df <- Iron %>% filter(oROI == s)
  # Try fitting the model
  model <- mgcv::gamm(
      tat2.combat ~ s(visit_age, k = 4, fx = FALSE, bs = "cs") + site,
      random = list(subject = ~1, subject = ~0 + visit_age),
      data = sub_df,
      method = "REML")
  
  # Run ANOVA on the GAM component
  aov_result <- anova(model$gam)
    F_val <- aov_result$pTerms.table[2]
    p_val <- aov_result$pTerms.table[3]
    
    # Append to results
    Iron_Results.site <- rbind(Iron_Results.site, data.frame(
      oROI = s,
      F_value = round(F_val, 2),
      p_value = signif(p_val, 2)
    ))
  }

Iron_Results.site$Label <- "Site"

#Peak Substance Use 

unique_peak_df <- dplyr::select(All_Sites_Iron, subject, sex, oIncome_recode, oRaceeth_recode, site, peak_Iron, peak_age_Iron,oROI) %>% filter(!is.na(oROI))
unique_peak_df <- unique(unique_peak_df)


# Across Substances 
gam.model_peakIron <- mgcv::gam(
  peak_Iron ~ oROI + sex + oIncome_recode + oRaceeth_recode + site,
  data = unique_peak_df %>% 
    filter(oROI != "BG"),
  method = "REML")
summary(gam.model_peakIron)
anova(gam.model_peakIron)  
 
library(mgcv)
library(dplyr)

# Initialize empty results data frame
Iron_Peak_Results.sex <- data.frame(
  oROI = character(),
  F_value = numeric(),
  p_value = numeric(),
  stringsAsFactors = FALSE
)

# Loop through each level of oSubstance
for (s in unique(unique_peak_df$oROI)) {
  
  # Subset data for the current substance
  sub_df <- unique_peak_df %>% filter(oROI == s)
  # Try fitting the model
  model <- mgcv::gam(
      peak_Iron ~ sex,
      data = sub_df,
      method = "REML")
  
  # Run ANOVA on the GAM component
  aov_result <- anova(model)
    F_val <- aov_result$pTerms.table[2]
    p_val <- aov_result$pTerms.table[3]
    
    # Append to results
    Iron_Peak_Results.sex <- rbind(Iron_Peak_Results.sex, data.frame(
      oROI = s,
      F_value = round(F_val, 2),
      p_value = signif(p_val, 2)
    ))
  }


Iron_Peak_Results.sex$Label <- "Sex"



# Initialize empty results data frame
Iron_Peak_Results.income <- data.frame(
  oROI = character(),
  F_value = numeric(),
  p_value = numeric(),
  stringsAsFactors = FALSE
)

# Loop through each level of oSubstance
for (s in unique(unique_peak_df$oROI)) {
  
  # Subset data for the current substance
  sub_df <- unique_peak_df %>% filter(oROI == s)
  # Try fitting the model
  model <- mgcv::gam(
      peak_Iron ~ oIncome_recode,
      data = sub_df,
      method = "REML")
  
  # Run ANOVA on the GAM component
  aov_result <- anova(model)
    F_val <- aov_result$pTerms.table[2]
    p_val <- aov_result$pTerms.table[3]
    
    # Append to results
    Iron_Peak_Results.income <- rbind(Iron_Peak_Results.income, data.frame(
      oROI = s,
      F_value = round(F_val, 2),
      p_value = signif(p_val, 2)
    ))
  }


Iron_Peak_Results.income$Label <- "Income"


# Initialize empty results data frame
Iron_Peak_Results.raceeth <- data.frame(
  oROI = character(),
  F_value = numeric(),
  p_value = numeric(),
  stringsAsFactors = FALSE
)

# Loop through each level of oSubstance
for (s in unique(unique_peak_df$oROI)) {
  
  # Subset data for the current substance
  sub_df <- unique_peak_df %>% filter(oROI == s)
  # Try fitting the model
  model <- mgcv::gam(
      peak_Iron ~ oRaceeth_recode,
      data = sub_df,
      method = "REML")
  
  # Run ANOVA on the GAM component
  aov_result <- anova(model)
    F_val <- aov_result$pTerms.table[2]
    p_val <- aov_result$pTerms.table[3]
    
    # Append to results
    Iron_Peak_Results.raceeth <- rbind(Iron_Peak_Results.raceeth, data.frame(
      oROI = s,
      F_value = round(F_val, 2),
      p_value = signif(p_val, 2)
    ))
  }


Iron_Peak_Results.raceeth$Label <- "Race/Eth"


# Initialize empty results data frame
Iron_Peak_Results.site <- data.frame(
  oROI = character(),
  F_value = numeric(),
  p_value = numeric(),
  stringsAsFactors = FALSE
)

# Loop through each level of oSubstance
for (s in unique(unique_peak_df$oROI)) {
  
  # Subset data for the current substance
  sub_df <- unique_peak_df %>% filter(oROI == s)
  # Try fitting the model
  model <- mgcv::gam(
      peak_Iron ~ site,
      data = sub_df,
      method = "REML")
  
  # Run ANOVA on the GAM component
  aov_result <- anova(model)
    F_val <- aov_result$pTerms.table[2]
    p_val <- aov_result$pTerms.table[3]
    
    # Append to results
    Iron_Peak_Results.site <- rbind(Iron_Peak_Results.site, data.frame(
      oROI = s,
      F_value = round(F_val, 2),
      p_value = signif(p_val, 2)
    ))
  }


Iron_Peak_Results.site$Label <- "Site"


#Combine models for data tables 
Main_Eff_Demo_Mean <- rbind(Iron_Results.sex, Iron_Results.income, Iron_Results.raceeth, Iron_Results.site)

custom_order <- c("BG", "Caudate", "NAcc", "Pallidum", "Putamen")

# Convert to factor with levels in custom order
Main_Eff_Demo_Mean <- Main_Eff_Demo_Mean %>%
  mutate(oROI = factor(oROI, levels = custom_order)) %>%
  arrange(oROI)

#Combine models for data tables 
Main_Eff_Demo_Peak <- rbind(Iron_Peak_Results.sex, Iron_Peak_Results.income, Iron_Peak_Results.raceeth, Iron_Peak_Results.site)

Main_Eff_Demo_Peak <- Main_Eff_Demo_Peak %>%
  mutate(oROI = factor(oROI, levels = custom_order)) %>%
  arrange(oROI)

Main_Eff_Demo_Mean[] <- lapply(Main_Eff_Demo_Mean, function(x) if (is.numeric(x)) round(x, 2) else x)
Main_Eff_Demo_Peak[] <- lapply(Main_Eff_Demo_Peak, function(x) if (is.numeric(x)) round(x, 2) else x)

write.csv(Main_Eff_Demo_Mean, file='/Users/ashparr/Library/Mobile Documents/com~apple~CloudDocs/Documents/Pitt/Manuscripts & Projects/NCANDA Project/DATA/Mean_Main_Effects_SociodemographicFacs_Across_IronROIs.csv', row.names = FALSE)

write.csv(Main_Eff_Demo_Peak, file='/Users/ashparr/Library/Mobile Documents/com~apple~CloudDocs/Documents/Pitt/Manuscripts & Projects/NCANDA Project/DATA/Peak_Main_Effects_SociodemographicFacs_Across_IronROIs.csv', row.names = FALSE)


``` 

### Plots: 
```{r, echo = FALSE, fig.align='center'}

#Order the demographic factors accordingly 
Long_means_all_SES$oLabel<-ordered(Long_means_all_SES$Label,
levels=c("E", "D", "C","B", "A", "Other","Non-Hispanic Black/AA", "Non-Hispanic White", "Hispanic", 
         "Asian AIAN NHPI OTHER", "200k+","100k-199k","75k-99k","50k-74k", "25k-49k","<25k", 
         "M","F","All"))

#Mean Iron across sociodemographic factors 
p1_5_SES <- Long_means_all_SES %>% ungroup () %>% 
  ggplot(aes(x = mean_Iron, y = oLabel, fill = oLabel)) + facet_wrap(~oROI, scales = "free")+
    scale_color_manual(values=Demo_Colors)+  
   scale_fill_manual(values=Demo_Colors)+  
  geom_point(shape = 21, size = 2.5, color = "white") +
labs(x = "\nnT2*w",
       y = "Subgroup\n") +
        # scale_x_continuous(trans = "reverse", limits = c(.015, .010), breaks = c(.015, .0125,.010))+ #Formatting best for consistency across all ROIs
scale_x_continuous(trans = "reverse", limits = c(.0139, .0133), breaks = c(.0138, .0136,.0134))+#Formatting best for basal ganglia iron (main)
    theme(aspect.ratio = 1.5,                       
        strip.background = element_blank(),
  legend.position = "none",
  axis.text = element_text(size = 8, family = "Arial", color = c("black")),
  axis.title.x = element_text(size = 9, family ="Arial", color = c("black")),
  axis.title.y = element_text(size = 9, family ="Arial", color = c("black")),
  axis.line = element_line(linewidth = .2), 
  axis.ticks = element_line(linewidth = .2),
  panel.grid.major = element_blank(),
  panel.grid.minor=element_blank(),
  panel.background=element_blank())
print(p1_5_SES)
 

#Mean Peak Iron across sociodemographic factors 
p2_5_SES <- Long_means_all_SES %>% ungroup () %>% 
  ggplot(aes(x = mean_peak_Iron, y = oLabel, fill = oLabel)) + facet_wrap(~oROI, scales = "free")+
      scale_color_manual(values=Demo_Colors)+  
   scale_fill_manual(values=Demo_Colors)+  
  geom_point(shape = 21, size = 2.5, color = "white") +
labs(x = "\nPeak nT2*w",
       y = "Subgroup\n") +
    # scale_x_continuous(trans = "reverse", limits = c(.015, .008), breaks = c(.014,.012, .010,.008))+ #Formatting best for consistency across all ROIs
scale_x_continuous(trans = "reverse", limits = c(.0133, .0129), breaks = c(.0133, .0131, .0129))+#Formatting best for basal ganglia iron (main)
  theme(aspect.ratio = 1.5,
 strip.background = element_blank(),
  legend.position = "none",
  axis.text = element_text(size = 8, family = "Arial", color = c("black")),
  axis.title.x = element_text(size = 9, family ="Arial", color = c("black")),
  axis.title.y = element_text(size = 9, family ="Arial", color = c("black")),
  axis.line = element_line(linewidth = .2), 
  axis.ticks = element_line(linewidth = .2),
  panel.grid.major = element_blank(),
  panel.grid.minor=element_blank(),
  panel.background=element_blank())
print(p2_5_SES)
 
#Mean Age of Peak Iron across sociodemographic factors 
p3_5_SES <- Long_means_all_SES %>% ungroup () %>% 
  ggplot(aes(x = mean_peak_age_Iron, y = oLabel, fill = oLabel)) + facet_wrap(~oROI, scales = "free")+
  geom_point(shape = 21, size = 2.5, color = "white") +
      scale_color_manual(values=Demo_Colors)+  
   scale_fill_manual(values=Demo_Colors)+  
theme_classic() +
  labs(x = "\nAge of Peak nT2*w",
       y = "Subgroup\n") +
          scale_x_continuous(limits = c(18.75, 22), breaks = seq(19, 22, by = 1))+ #Formatting best for general substance use (all)
        # scale_x_continuous(limits = c(18.75, 22), breaks = seq(19, 22, by = 1))+ #Formatting best for consistency across all substances
theme(aspect.ratio = 1.5,
  strip.background = element_blank(),
  legend.position = "none",
  axis.text = element_text(size = 8, family = "Arial", color = c("black")),
  axis.title.x = element_text(size = 9, family ="Arial", color = c("black")),
  axis.title.y = element_text(size = 9, family ="Arial", color = c("black")),
  axis.line = element_line(linewidth = .2), 
  axis.ticks = element_line(linewidth = .2),
  panel.grid.major = element_blank(),
  panel.grid.minor=element_blank(),
  panel.background=element_blank())
print(p3_5_SES)
 

ggsave(p1_5_SES, filename = "/Users/ashparr/Documents/Pitt/Manuscripts & Projects/NCANDA Project/Results 2025/Iron_Mean_AllROIs_SES.pdf", device = "pdf", dpi = 500,
       width = 8, height = 8, units = "in")

ggsave(p2_5_SES, filename = "/Users/ashparr/Documents/Pitt/Manuscripts & Projects/NCANDA Project/Results 2025/Iron_Peak_AllROIs_SES.pdf", device = "pdf", dpi = 500,
       width = 8, height = 8, units = "in")
 
ggsave(p3_5_SES, filename = "/Users/ashparr/Documents/Pitt/Manuscripts & Projects/NCANDA Project/Results 2025/Iron_Peak_Age_AllROIs_SES.pdf", device = "pdf", dpi = 500,
       width = 8, height = 8, units = "in")
  
``` 

# Characterize Age Changes
## Full Sample   

```{r, echo = FALSE, fig.align='center'}

 
rois <- levels(Iron$roi)

gam.statistics <- data.frame(
  roi = factor(rep(NA, length(rois)), levels = rois),
  GAM.smooth_Fvalue = numeric(length(rois)),
  GAM.smooth.pvalue = numeric(length(rois)),
  smooth.change.onset = numeric(length(rois)),
  smooth.peak.change = numeric(length(rois)),
  smooth.decrease.onset = numeric(length(rois)),
  smooth.decrease.offset = numeric(length(rois)),
  smooth.increase.onset = numeric(length(rois)),
  smooth.increase.offset = numeric(length(rois)),
  smooth.last.change = numeric(length(rois)), 
  max.age.fitted = numeric(length(rois))
  )

#Substance = roi
gam.fittedvalues <- data.frame(roi = factor(), visit_age = numeric(), fitted = numeric(), fitted2 = numeric(), se = numeric(), lower_ci = numeric(), upper_ci=numeric())

gam.smoothestimates <- data.frame(roi = factor(), visit_age = numeric(), estimate = numeric(), se = numeric())

gam.derivatives <- data.frame(roi = factor(), visit_age = numeric(), derivative = numeric(), lower = numeric(),
                              upper = numeric(), significant = numeric(), significant_deriv = numeric()) 

gam.pred <- data.frame(roi = factor(), visit_age = numeric(), predicted = numeric(), se = numeric(),
                              lower = numeric(), upper = numeric()) 

#subst = roist
for (roist in seq_along(rois)) {
  sub_df <- Iron[Iron$roi == rois[roist], ]  # Subset data
  
gam.model <- mgcv::gamm(tat2.combat ~ s(visit_age, fx = F, k = 4, bs = "cs") + 
                              sex + oIncome_recode + oRaceeth_recode + site, 
             random = list(subject = ~1, subject = ~0 + visit_age),
             data=sub_df,# %>%
             method = "REML")

gam.results <- summary(gam.model$gam)

df <- as.data.frame(unclass(gam.model$gam$model),stringsAsFactors = TRUE)
np <- 200
thisPred <- data.frame(init = rep(0,np))
theseVars <- attr(gam.model$gam$terms, "term.labels")
varClasses <- attr(gam.model$gam$terms,"dataClasses")
thisResp <- as.character(gam.model$gam$terms[[2]])
smooth_var <- "visit_age"
for (v in c(1:length(theseVars))) {
  thisVar <- theseVars[[v]]
  thisClass <- varClasses[thisVar]
  if (thisVar == smooth_var) {
    thisPred[,smooth_var] = seq(min(df[,smooth_var],na.rm=T), max(df[,smooth_var], na.rm=T), length.out = np)
  } else {
      switch (thisClass,
              "numeric" = {thisPred[,thisVar] = median(df[,thisVar])}, #make predictions based on median value
              "character" = {thisPred[,thisVar] = levels(df[,thisVar])[[1]]}, #make predictions based on first level of char
              "factor" = {thisPred[,thisVar] = levels(df[,thisVar])[[1]]}, #make predictions based on first level of factor 
              "ordered" = {thisPred[,thisVar] = levels(df[,thisVar])[[1]]} #make predictions based on first level of ordinal variable
      )
    }
  }
  pred <- thisPred %>% dplyr::select(-init)
  
## MODEL STATSTICS ##  
#GAM derivatives
#Get derivatives of the smooth function using finite differences
derv <- gratia::derivatives(gam.model, term = sprintf('s(%s)',smooth_var), data = pred, interval = "simultaneous", unconditional = F) #derivative at 200 indices of smooth_var with a simultaneous CI
  #Identify derivative significance window(s)
  derv <- derv %>% #add "sig" column (TRUE/FALSE) to derv
    mutate(sig = !(0 > .lower_ci & 0 < .upper_ci)) #derivative is sig if the lower CI is not < 0 while the upper CI is > 0 (i.e., when the CI does not include 0)
  derv$sig_deriv = derv$.derivative*derv$sig #add "sig_deriv derivatives column where non-significant derivatives are set to 0
  mean.derivative <- mean(derv$.derivative)
  
  #Model summary outputs
  #Signed F value for the smooth term and GAM-based significance of the smooth term
  gam.smooth.F <- gam.results$s.table[1,3] #first row is s(smooth_var), third entry is F
  gam.smooth.Rsq <- gam.results$r.sq
  if(mean.derivative < 0){ #if the average derivative is less than 0, make the effect size estimate negative
    gam.smooth.F <- gam.smooth.F*-1}
  
  gam.smooth.pvalue <- gam.results$s.table[1,4] #first row is s(smooth_var), fourth entry is p-value

 #Derivative-based temporal characteristics
  #Age of developmental change onset
  if(sum(derv$sig) > 0){ #if derivative is significant at at least 1 age
    change.onset <- min(derv$visit_age[derv$sig==T])} #find first age in the smooth where derivative is significant
  if(sum(derv$sig) == 0){ #if gam derivative is never significant
    change.onset <- NA} #assign NA
 
  #Age of maximal developmental change
  if(sum(derv$sig) > 0){ 
    derv$abs_sig_deriv = round(abs(derv$sig_deriv),5) #absolute value significant derivatives
    maxval <- max(derv$abs_sig_deriv) #find the largest derivative
    window.peak.change <- derv$visit_age[derv$abs_sig_deriv == maxval] #identify the age(s) at which the derivative is greatest in absolute magnitude
    peak.change <- mean(window.peak.change)} #identify the age of peak developmental change
  if(sum(derv$sig) == 0){ 
    peak.change <- NA}  
 
  #Age of decrease onset
  if(sum(derv$sig) > 0){ 
    decreasing.range <- derv$visit_age[derv$sig_deriv < 0] #identify all ages with a significant negative derivative (i.e., smooth_var indices where y is decreasing)
    if(length(decreasing.range) > 0)
      decrease.onset <- min(decreasing.range) #find youngest age with a significant negative derivative
    if(length(decreasing.range) == 0)
      decrease.onset <- NA}
  if(sum(derv$sig) == 0){
    decrease.onset <- NA}  

    #Age of decrease offset
  if(sum(derv$sig) > 0){ 
    decreasing.range <- derv$visit_age[derv$sig_deriv < 0] #identify all ages with a significant negative derivative (i.e., smooth_var indices where y is decreasing)
    if(length(decreasing.range) > 0){
      last.decrease <- max(decreasing.range) #find oldest age with a significant negative derivative
      if(last.decrease == derv$visit_age[length(derv$visit_age)]) #if the last age of significant decrease is the oldest in the dataset
        decrease.offset <- last.decrease
      if(last.decrease != derv$visit_age[length(derv$visit_age)]){ 
        decrease.offset.row <- which(derv$visit_age == last.decrease) + 1 #use above to find the first age when the derivative is not significant
        decrease.offset <- derv$visit_age[decrease.offset.row]}
    }
    if(length(decreasing.range) == 0)
      decrease.offset <- NA}
  if(sum(derv$sig) == 0){ 
    decrease.offset <- NA}  

  #Age of increase onset
  if(sum(derv$sig) > 0){ 
    increasing.range <- derv$visit_age[derv$sig_deriv > 0] #identify all ages with a significant positive derivative (i.e., smooth_var indices where y is increasing)
    if(length(increasing.range) > 0)
      increase.onset <- min(increasing.range) #find oldest age with a significant positive derivative
    if(length(increasing.range) == 0)
      increase.onset <- NA}
  if(sum(derv$sig) == 0){ 
    increase.onset <- NA}  
 
#Age of increase offset
  if(sum(derv$sig) > 0){ 
    increasing.range <- derv$visit_age[derv$sig_deriv > 0] #identify all ages with a significant positive derivative (i.e., smooth_var indices where y is increasing)
    if(length(increasing.range) > 0){
      last.increase <- max(increasing.range) #find oldest age with a significant positive derivative
      if(last.increase == derv$visit_age[length(derv$visit_age)]) #if the last age of significant increase is the oldest in the dataset
        increase.offset <- last.increase
      if(last.increase != derv$visit_age[length(derv$visit_age)]){ 
        increase.offset.row <- which(derv$visit_age == last.increase) + 1 #use above to find the first age when the derivative is not significant
        increase.offset <- derv$visit_age[increase.offset.row]}
    }
    if(length(increasing.range) == 0)
      increase.offset <- NA}
  if(sum(derv$sig) == 0){ 
    increase.offset <- NA}  
  
 #Age of last change
  if(sum(derv$sig) > 0){ 
    change.offset <- max(derv$visit_age[derv$sig==T])} #find last age in the smooth where derivative is significant
  if(sum(derv$sig) == 0){ 
    change.offset <- NA}  
  
gam.statistics$roi[roist] <- rois[roist]
gam.statistics$GAM.smooth_Fvalue[roist] <- gam.smooth.F
gam.statistics$GAM.smooth_RSq[roist] <- gam.smooth.Rsq
gam.statistics$GAM.smooth.pvalue[roist] <- gam.smooth.pvalue
gam.statistics$smooth.change.onset[roist] <- change.onset
gam.statistics$smooth.peak.change[roist] <- peak.change
gam.statistics$smooth.decrease.onset[roist] <- decrease.onset
gam.statistics$smooth.decrease.offset[roist] <- decrease.offset
gam.statistics$smooth.increase.onset[roist] <- increase.onset
gam.statistics$smooth.increase.offset[roist] <- increase.offset
gam.statistics$smooth.last.change[roist] <- change.offset

 #Generate predictions (fitted values) based on the gam model and predication data frame
gam.fittedvalues_temp <- gratia::fitted_values(object = gam.model, data = pred)
gam.fittedvalues_temp$.fitted2 <- plogis(gam.fittedvalues_temp$.fitted) * 30
  
maxval2 <- min(gam.fittedvalues_temp$.fitted) #find the largest derivative
gam.statistics$max.age.fitted[roist] <- gam.fittedvalues_temp$visit_age[gam.fittedvalues_temp$.fitted == maxval2] #identify the age(s) at which the 
gam.statistics$max.val.fitted[roist] <- maxval2

# gam.fittedvalues_temp <- plogis(fitted(gam.model$gam))*30

temp_fit <- data.frame(
  roi = factor(rois[roist]),
  visit_age = gam.fittedvalues_temp$visit_age, 
  fitted = gam.fittedvalues_temp$.fitted, 
  fitted2 = gam.fittedvalues_temp$.fitted2, 
  se = gam.fittedvalues_temp$.se, 
  lower_ci = gam.fittedvalues_temp$.lower_ci, 
  upper_ci = gam.fittedvalues_temp$.upper_ci)
gam.fittedvalues <- bind_rows(gam.fittedvalues, temp_fit)

gam.smoothvalues_temp <- gratia::smooth_estimates(object = gam.model, data = pred, smooth = sprintf('s(%s)',smooth_var))

temp_smooth <- data.frame(
  roi = factor(rois[roist]),
  visit_age = gam.smoothvalues_temp$visit_age, 
  estimate = gam.smoothvalues_temp$.estimate,  
  se = gam.smoothvalues_temp$.se)
gam.smoothestimates <- bind_rows(gam.smoothestimates, temp_smooth)

temp_deriv <- data.frame(
  roi = factor(rois[roist]),
  visit_age = derv$visit_age, 
  derivative = derv$.derivative, 
  lower = derv$.lower_ci,
  upper = derv$.upper_ci, 
  significant = derv$sig, 
  significant_deriv = derv$sig_deriv)

gam.derivatives <- bind_rows(gam.derivatives, temp_deriv)

gam.pred_temp <- ggpredict(gam.model, terms=c("visit_age"))

temp_pred <- data.frame(
  roi = factor(rois[roist]),
  visit_age = gam.pred_temp$x,
  predicted = gam.pred_temp$predicted, 
  se = gam.pred_temp$std.error,
  lower = gam.pred_temp$conf.low, 
  upper = gam.pred_temp$conf.high)

gam.pred <- bind_rows(gam.pred, temp_pred)  

}


#Calculate the average first derivative in each region at each substance
deriv_rate <- gam.derivatives %>% group_by(roi) %>% do(deriv_rate = mean(.$derivative)) %>% unnest(cols = "deriv_rate")

gam.statistics <- merge(gam.statistics, deriv_rate, by = "roi")



# gam.statistics[] <- lapply(gam.statistics, function(x) if (is.numeric(x)) round(x, 2) else x)

custom_order <- c("BG", "Caudate", "NAcc", "Pallidum", "Putamen")
# Convert to factor with levels in custom order
gam.statistics <- gam.statistics %>%
  mutate(roi = factor(roi, levels = custom_order)) %>%
  arrange(roi)

write.csv(gam.statistics, file='/Users/ashparr/Library/Mobile Documents/com~apple~CloudDocs/Documents/Pitt/Manuscripts & Projects/NCANDA Project/DATA/Smooth_Statistics_Across_IronROIs_FullSample.csv', row.names = FALSE)


``` 

### Plots:  
```{r, echo = FALSE, fig.align='center'}

# Age by Iron: All Subscsales, no datapoints
age1 <- ggplot(All_Sites_Iron,
                    aes(x=visit_age, y=tat2.combat, color=oROI,fill=oROI))+
  geom_smooth(method="gam", formula = y ~ s(x, k = 4, fx = FALSE, bs = "cs"))+
   scale_color_manual(values=iron_colorbar)+
   scale_fill_manual(values=iron_colorbar)+
   theme(aspect.ratio=.5, legend.position = "none",
         axis.text = element_text(size = 8, family = "Arial", color = c("black")),
  axis.title.x = element_text(size = 9, family ="Arial", color = c("black")),
  axis.title.y = element_text(size = 9, family ="Arial", color = c("black")),
  axis.line = element_line(linewidth = .2), 
  axis.ticks = element_line(linewidth = .2),
        panel.grid.major = element_blank(),panel.grid.minor=element_blank(),
        panel.background=element_blank()) +
  scale_x_continuous(limits = c(12, 30), breaks = seq(15, 30, by = 5))+
  scale_y_continuous(trans = "reverse")+
   labs(x = "\nAge",
       y = "nT2*w\n")
print(age1)

# Age by Iron: ROIs separately, trajectories shown 
age2 <- ggplot(All_Sites_Iron,
                    aes(x=visit_age, y=tat2.combat, color=oROI,fill=oROI))+ 
  geom_line(data= All_Sites_Iron,
            aes(y=tat2.combat, x = visit_age, group=subject), alpha=.1, color = "black",
            linewidth = .25)+ facet_wrap(~oROI, scales = "free")+
  scale_fill_manual(values = iron_colorbar)+ 
  scale_color_manual(values = iron_colorbar)+ 
  geom_smooth(method="gam", formula = y ~ s(x, k = 4, fx = FALSE,bs = "cs"))+
  theme(legend.position = "none", aspect.ratio=.5, 
                               strip.background = element_blank(),
         axis.text = element_text(size = 8, family = "Arial", color = c("black")),
  axis.title.x = element_text(size = 9, family ="Arial", color = c("black")),
  axis.title.y = element_text(size = 9, family ="Arial", color = c("black")),
  axis.line = element_line(linewidth = .2), 
  axis.ticks = element_line(linewidth = .2),
        panel.grid.major = element_blank(),panel.grid.minor=element_blank(),
        panel.background=element_blank()) +
  scale_x_continuous(limits = c(12, 30), breaks = seq(15, 30, by = 5))+
  scale_y_continuous(trans = "reverse")+
  scale_y_continuous(trans = "reverse",limits = c(.016, .011))+ #Formatting best for basal ganglia (main)
    # scale_y_continuous(trans = "reverse",limits = c(.02, .003))+#Formatting best for consistency across ROIs
    labs(x = "\nAge",
       y = "nT2*w\n")
print(age2)
 
# Derivative plots for Iron: ROIs separately 

derv<- gam.derivatives[gam.derivatives$roi=="BG",] 
d1<- ggplot(data=derv) + geom_tile(aes(x = visit_age, y = .5, fill = significant_deriv))+
      scale_fill_gradient2(low = iron_colorbar[1], midpoint = 0, mid = "white",
                           high = iron_colorbar[1])+#, limits = c(min_val, 0))+ 
    theme(aspect.ratio = .05, panel.grid.major = element_blank(),
          panel.grid.minor=element_blank(),
          panel.background=element_blank(), 
          axis.text = element_text(size = 8, family = "Arial", color = c("black")),
          axis.title.x = element_text(size = 9, family ="Arial", color = c("black")),
          axis.title.y = element_text(size = 9, family ="Arial", color = c("black")),
          axis.line = element_line(linewidth = .2), 
          axis.ticks = element_line(linewidth = .2),
          legend.key.width = unit(1,"cm"),
          legend.key.height = unit(.5,"cm"),
          legend.position = "bottom",
          plot.margin = unit(c(0, 0, 0, 0), "cm"))+
    guides(fill = guide_colorbar(reverse = F,direction = "horizontal",title.position = "top")) + geom_vline(data = gam.statistics[gam.statistics$roi=="BG",], aes(xintercept=smooth.decrease.offset), color="black") +
      geom_vline(data = gam.statistics[gam.statistics$roi=="BG",], aes(xintercept=smooth.decrease.onset), color="black") +
scale_x_continuous(limits = c(12, 30), breaks = seq(15, 30, by = 5))+
  xlab("\nAge") + ylab("Basal Ganglia nT2*w\n")
 
derv<- gam.derivatives[gam.derivatives$roi=="NAcc",] 
d2<- ggplot(data=derv) + geom_tile(aes(x = visit_age, y = .5, fill = significant_deriv))+
      scale_fill_gradient2(low = iron_colorbar[2], midpoint = 0, mid = "white",
                           high = iron_colorbar[2])+#, limits = c(0, max_val))+ 
    theme(aspect.ratio = .05, panel.grid.major = element_blank(),
          panel.grid.minor=element_blank(),
          panel.background=element_blank(), 
          axis.text = element_text(size = 8, family = "Arial", color = c("black")),
          axis.title.x = element_text(size = 9, family ="Arial", color = c("black")),
          axis.title.y = element_text(size = 9, family ="Arial", color = c("black")),
          axis.line = element_line(linewidth = .2), 
          axis.ticks = element_line(linewidth = .2),
          legend.key.width = unit(1,"cm"),
          legend.key.height = unit(.5,"cm"),
          legend.position = "bottom",
          plot.margin = unit(c(0, 0, 0, 0), "cm"))+
    guides(fill = guide_colorbar(reverse = F,direction = "horizontal",title.position = "top")) + 
  geom_vline(data = gam.statistics[gam.statistics$roi=="NAcc",], aes(xintercept=smooth.decrease.offset), color="black") +
      geom_vline(data = gam.statistics[gam.statistics$roi=="NAcc",], aes(xintercept=smooth.decrease.onset), color="black") +
  scale_x_continuous(limits = c(12, 30), breaks = seq(15, 30, by = 5))+  xlab("\nAge") + ylab("NAcc nT2*w\n")

derv<- gam.derivatives[gam.derivatives$roi=="Putamen",] 
d3<- ggplot(data=derv) + geom_tile(aes(x = visit_age, y = .5, fill = significant_deriv))+
      scale_fill_gradient2(low = iron_colorbar[3], midpoint = 0, mid = "white",
                           high = iron_colorbar[3])+#, limits = c(0, max_val))+ 
    theme(aspect.ratio = .05, panel.grid.major = element_blank(),
          panel.grid.minor=element_blank(),
          panel.background=element_blank(), 
          axis.text = element_text(size = 8, family = "Arial", color = c("black")),
          axis.title.x = element_text(size = 9, family ="Arial", color = c("black")),
          axis.title.y = element_text(size = 9, family ="Arial", color = c("black")),
          axis.line = element_line(linewidth = .2), 
          axis.ticks = element_line(linewidth = .2),
          legend.key.width = unit(1,"cm"),
          legend.key.height = unit(.5,"cm"),
          legend.position = "bottom",
          plot.margin = unit(c(0, 0, 0, 0), "cm"))+
    guides(fill = guide_colorbar(reverse = F,direction = "horizontal",title.position = "top")) +   geom_vline(data = gam.statistics[gam.statistics$roi=="Putamen",], aes(xintercept=smooth.decrease.offset), color="black") +
      geom_vline(data = gam.statistics[gam.statistics$roi=="Putamen",], aes(xintercept=smooth.decrease.onset), color="black") +
  scale_x_continuous(limits = c(12, 30), breaks = seq(15, 30, by = 5))+  xlab("\nAge") + ylab("Putamen nT2*w\n")


derv<- gam.derivatives[gam.derivatives$roi=="Caudate",] 
d4<- ggplot(data=derv) + geom_tile(aes(x = visit_age, y = .5, fill = significant_deriv))+
      scale_fill_gradient2(low = iron_colorbar[4], midpoint = 0, mid = "white",
                           high = iron_colorbar[4])+#, limits = c(0, max_val))+ 
    theme(aspect.ratio = .05, panel.grid.major = element_blank(),
          panel.grid.minor=element_blank(),
          panel.background=element_blank(), 
          axis.text = element_text(size = 8, family = "Arial", color = c("black")),
          axis.title.x = element_text(size = 9, family ="Arial", color = c("black")),
          axis.title.y = element_text(size = 9, family ="Arial", color = c("black")),
          axis.line = element_line(linewidth = .2), 
          axis.ticks = element_line(linewidth = .2),
          legend.key.width = unit(1,"cm"),
          legend.key.height = unit(.5,"cm"),
          legend.position = "bottom",
          plot.margin = unit(c(0, 0, 0, 0), "cm"))+
    guides(fill = guide_colorbar(reverse = F,direction = "horizontal",title.position = "top")) +   geom_vline(data = gam.statistics[gam.statistics$roi=="Caudate",], aes(xintercept=smooth.decrease.offset), color="black") +
      geom_vline(data = gam.statistics[gam.statistics$roi=="Caudate",], aes(xintercept=smooth.decrease.onset), color="black") +
  scale_x_continuous(limits = c(12, 30), breaks = seq(15, 30, by = 5))+  xlab("\nAge") + ylab("Caudate nT2*w\n")
 
derv<- gam.derivatives[gam.derivatives$roi=="Pallidum",] 
d5<- ggplot(data=derv) + geom_tile(aes(x = visit_age, y = .5, fill = significant_deriv))+
      scale_fill_gradient2(low = iron_colorbar[5], midpoint = 0, mid = "white",
                           high = iron_colorbar[5])+#, limits = c(0, max_val))+ 
    theme(aspect.ratio = .05, panel.grid.major = element_blank(),
          panel.grid.minor=element_blank(),
          panel.background=element_blank(), 
          axis.text = element_text(size = 8, family = "Arial", color = c("black")),
          axis.title.x = element_text(size = 9, family ="Arial", color = c("black")),
          axis.title.y = element_text(size = 9, family ="Arial", color = c("black")),
          axis.line = element_line(linewidth = .2), 
          axis.ticks = element_line(linewidth = .2),
          legend.key.width = unit(1,"cm"),
          legend.key.height = unit(.5,"cm"),
          legend.position = "bottom",
          plot.margin = unit(c(0, 0, 0, 0), "cm"))+
    guides(fill = guide_colorbar(reverse = F,direction = "horizontal",title.position = "top")) +   geom_vline(data = gam.statistics[gam.statistics$roi=="Pallidum",], aes(xintercept=smooth.decrease.offset), color="black") +
      geom_vline(data = gam.statistics[gam.statistics$roi=="Pallidum",], aes(xintercept=smooth.decrease.onset), color="black") +
  scale_x_continuous(limits = c(12, 30), breaks = seq(15, 30, by = 5))+  xlab("\nAge") + ylab("Pallidum nT2*w\n")




Derv1 <- ggarrange(d1, d2, d3, 
                   ncol = 3, nrow = 1)
  
Derv2 <- ggarrange(d4, d5, 
                   ncol = 3, nrow = 1)
 

ggsave(age1, filename = "/Users/ashparr/Documents/Pitt/Manuscripts & Projects/NCANDA Project/Results 2025/Iron_Age_AllROIs.pdf", device = "pdf", dpi = 500,
       width = 8, height = 8, units = "in")

ggsave(age2, filename = "/Users/ashparr/Documents/Pitt/Manuscripts & Projects/NCANDA Project/Results 2025/Iron_Age_AllROIs_Sep.pdf", device = "pdf", dpi = 500,
       width = 8, height = 8, units = "in")

ggsave(Derv1, filename = "/Users/ashparr/Documents/Pitt/Manuscripts & Projects/NCANDA Project/Results 2025/Iron_Age_Deriv1_AllROIs_Sep.pdf", device = "pdf", dpi = 500,
       width = 8, height = 8, units = "in")


ggsave(Derv2, filename = "/Users/ashparr/Documents/Pitt/Manuscripts & Projects/NCANDA Project/Results 2025/Iron_Age_Deriv2_AllROIs_Sep.pdf", device = "pdf", dpi = 500,
       width = 8, height = 8, units = "in")



gam.statistics$oROI <- ordered(gam.statistics$roi, levels = c( "BG","NAcc", "Pallidum","Putamen", 
                                                                          "Caudate"))

# Age of Peak Iron Identified by smooth fits in GAM models

age1_5 <- gam.statistics %>% ungroup () %>%  
  ggplot(aes(x = max.age.fitted, y = oROI, fill = oROI)) +
  geom_point(shape = 21, size = 5.5, color = "white") +
  scale_fill_manual(values = iron_colorbar2) +
labs(x = "\nAge of Peak nT2*w (smooth)",
       y = "ROI\n") +
  scale_y_discrete(
    breaks = c("BG", "NAcc", "Pallidum", "Putamen", "Caudate"),
    labels = c("BG", "NAcc", "GP","Put.","Caud."))+
            scale_x_continuous(limits = c(29, 29.5), breaks = seq(29, 29.5, by = .25))+
  theme(aspect.ratio = 1.5,
  legend.position = "none",
  axis.text = element_text(size = 8, family = "Arial", color = c("black")),
  axis.title.x = element_text(size = 9, family ="Arial", color = c("black")),
  axis.title.y = element_text(size = 9, family ="Arial", color = c("black")),
  axis.line = element_line(linewidth = .2), 
  axis.ticks = element_line(linewidth = .2),
  panel.grid.major = element_blank(),
  panel.grid.minor=element_blank(),
  panel.background=element_blank())
print(age1_5)

age2_5 <- gam.statistics %>% ungroup () %>% 
  ggplot(aes(x = deriv_rate, y = oROI, fill = oROI)) +
  geom_point(shape = 21, size = 5.5, color = "white") +
  scale_fill_manual(values = iron_colorbar2) +
labs(x = "\nRate of Increase",
       y = "ROI\n") +
  scale_y_discrete(
    breaks = c("BG", "NAcc", "Pallidum", "Putamen", "Caudate"),
    labels = c("BG", "NAcc", "GP","Put.","Caud."))+
    scale_x_continuous(trans = "reverse")+
      # scale_x_continuous(limits = c(0, .25))+
  # scale_x_continuous(breaks = c(0.0010, 0.0014, 0.0018), limits = c(0.00085, 0.0019)) +
  theme(aspect.ratio = 1.5,
  legend.position = "none",
  axis.text = element_text(size = 8, family = "Arial", color = c("black")),
  axis.title.x = element_text(size = 9, family ="Arial", color = c("black")),
  axis.title.y = element_text(size = 9, family ="Arial", color = c("black")),
  axis.line = element_line(linewidth = .2), 
  axis.ticks = element_line(linewidth = .2),
  panel.grid.major = element_blank(),
  panel.grid.minor=element_blank(),
  panel.background=element_blank())
print(age2_5)



age3_5 <- gam.statistics %>% ungroup () %>% 
  ggplot(aes(x = smooth.last.change, y = oROI, fill = oROI)) +
  geom_point(shape = 21, size = 5.5, color = "white") +
  scale_fill_manual(values = iron_colorbar2) +
labs(x = "\nAge of Maturation",
       y = "ROI\n") +
  scale_y_discrete(
    breaks = c("BG", "NAcc", "Pallidum", "Putamen", "Caudate"),
    labels = c("BG", "NAcc", "GP","Put.","Caud."))+
            scale_x_continuous(limits = c(29, 29.5), breaks = seq(29, 29.5, by = .25))+
  theme(aspect.ratio = 1.5,
  legend.position = "none",
  axis.text = element_text(size = 8, family = "Arial", color = c("black")),
  axis.title.x = element_text(size = 9, family ="Arial", color = c("black")),
  axis.title.y = element_text(size = 9, family ="Arial", color = c("black")),
  axis.line = element_line(linewidth = .2), 
  axis.ticks = element_line(linewidth = .2),
  panel.grid.major = element_blank(),
  panel.grid.minor=element_blank(),
  panel.background=element_blank())
print(age3_5)


age4_5 <- gam.statistics %>% ungroup () %>% 
  ggplot(aes(x = smooth.peak.change, y = oROI, fill = oROI)) +
  geom_point(shape = 21, size = 5.5, color = "white") +
  scale_fill_manual(values = iron_colorbar2) +
 labs(x = "\nAge of Peak Change",
       y = "ROI\n") +
  scale_y_discrete(
    breaks = c("BG", "NAcc", "Pallidum", "Putamen", "Caudate"),
    labels = c("BG", "NAcc", "GP","Put.","Caud."))+
scale_x_continuous(limits = c(12, 30), breaks = seq(15, 30, by = 5))+
  theme(aspect.ratio = 1.5,
  legend.position = "none",
  axis.text = element_text(size = 8, family = "Arial", color = c("black")),
  axis.title.x = element_text(size = 9, family ="Arial", color = c("black")),
  axis.title.y = element_text(size = 9, family ="Arial", color = c("black")),
  axis.line = element_line(linewidth = .2), 
  axis.ticks = element_line(linewidth = .2),
  panel.grid.major = element_blank(),
  panel.grid.minor=element_blank(),
  panel.background=element_blank())
print(age4_5)



age5_5 <- gam.statistics %>% ungroup () %>% 
  ggplot(aes(x = smooth.decrease.onset, y = oROI, fill = oROI)) +
  geom_point(shape = 21, size = 5.5, color = "white") +

  scale_fill_manual(values = iron_colorbar2) +
  labs(x = "\nIncrease Onset (Age)",
       y = "ROI\n") +
  scale_y_discrete(
    breaks = c("BG", "NAcc", "Pallidum", "Putamen", "Caudate"),
    labels = c("BG", "NAcc", "GP","Put.","Caud."))+
            scale_x_continuous(limits = c(12, 15), breaks = seq(12, 15, by = 1))+
  theme(aspect.ratio = 1.5,
  legend.position = "none",
  axis.text = element_text(size = 8, family = "Arial", color = c("black")),
  axis.title.x = element_text(size = 9, family ="Arial", color = c("black")),
  axis.title.y = element_text(size = 9, family ="Arial", color = c("black")),
  axis.line = element_line(linewidth = .2), 
  axis.ticks = element_line(linewidth = .2),
  panel.grid.major = element_blank(),
  panel.grid.minor=element_blank(),
  panel.background=element_blank())
print(age5_5)



age6_5 <- gam.statistics %>% ungroup () %>%  
  ggplot(aes(x = smooth.decrease.offset, y = oROI, fill = oROI)) +
  geom_point(shape = 21, size = 5.5, color = "white") +
  scale_fill_manual(values = iron_colorbar2) +
  labs(x = "\nIncrease Offset (Age)",
       y = "ROI\n") +
  scale_y_discrete(
    breaks = c("BG", "NAcc", "Pallidum", "Putamen", "Caudate"),
    labels = c("BG", "NAcc", "GP","Put.","Caud."))+
            scale_x_continuous(limits = c(29, 29.5), breaks = seq(29, 29.5, by = .25))+
  theme(aspect.ratio = 1.5,
  legend.position = "none",
  axis.text = element_text(size = 8, family = "Arial", color = c("black")),
  axis.title.x = element_text(size = 9, family ="Arial", color = c("black")),
  axis.title.y = element_text(size = 9, family ="Arial", color = c("black")),
  axis.line = element_line(linewidth = .2), 
  axis.ticks = element_line(linewidth = .2),
  panel.grid.major = element_blank(),
  panel.grid.minor=element_blank(),
  panel.background=element_blank())
print(age6_5)


Iron_Plots_Age1 <- ggarrange(age1_5, age2_5, age3_5,
                    ncol = 3, nrow = 1)
Iron_Plots_Age1
 

Iron_Plots_Age2 <- ggarrange(age4_5, age5_5, age6_5,
                    ncol = 3, nrow = 1)
Iron_Plots_Age2 

ggsave(Iron_Plots_Age1, filename = "/Users/ashparr/Documents/Pitt/Manuscripts & Projects/NCANDA Project/Results 2025/Iron_Age_AllROIs_Gam_Stats1.pdf", device = "pdf", dpi = 500,
       width = 8, height = 8, units = "in")#

ggsave(Iron_Plots_Age2, filename = "/Users/ashparr/Documents/Pitt/Manuscripts & Projects/NCANDA Project/Results 2025/Iron_Age_AllROIs_Gam_Stats2.pdf", device = "pdf", dpi = 500,
       width = 8, height = 8, units = "in")#


``` 
## Demographic Factors   

### All Participants (no covariates)
```{r, echo = FALSE, fig.align='center'}

set.seed(1)


roi <- levels(Iron$oROI)

# Initialize output containers
gam.statistics <- data.frame(
  oROI = factor(),
  Label = factor(),
  GAM.smooth_Fvalue = numeric(),
  GAM.smooth_RSq = numeric(),
  GAM.smooth.pvalue = numeric(),
  smooth.change.onset = numeric(),
  smooth.peak.change = numeric(),
  smooth.decrease.onset = numeric(),
  smooth.decrease.offset = numeric(),
  smooth.increase.onset = numeric(),
  smooth.increase.offset = numeric(),
  smooth.last.change = numeric(),
  max.age.fitted = numeric(),
  max.val.fitted = numeric()
)

gam.fittedvalues <- data.frame(
  oROI = factor(),
  Label = factor(),
  visit_age = numeric(),
  fitted = numeric(),
  fitted2 = numeric(),
  se = numeric(),
  lower_ci = numeric(),
  upper_ci = numeric()
)

gam.smoothestimates <- data.frame(
  oROI = factor(),
  Label = factor(),
  visit_age = numeric(),
  estimate = numeric(),
  se = numeric()
)

gam.derivatives <- data.frame(
  oROI = factor(),
  Label = factor(),
  visit_age = numeric(),
  derivative = numeric(),
  lower = numeric(),
  upper = numeric(),
  significant = numeric(),
  significant_deriv = numeric()
)

gam.pred <- data.frame(
  oROI = factor(),
  Label = factor(),
  visit_age = numeric(),
  predicted = numeric(),
  se = numeric(),
  lower = numeric(),
  upper = numeric()
)

# ----------------------------
# Loop over ROI & sexes
# ----------------------------
for (subst in roi) {
    
    sub_df <- Iron %>%
      filter(oROI == subst)
    
    # if (nrow(sub_df) < 10) next  # skip if too few rows
    
    gam.model <- mgcv::gamm(
      tat2.combat ~ s(visit_age, k = 4, fx = F, bs = "cs"),
      random = list(subject = ~1, subject = ~0 + visit_age),
      data = sub_df,
      method = "REML"
    )
    
    gam.results <- summary(gam.model$gam)
    df <- as.data.frame(unclass(gam.model$gam$model), stringsAsFactors = TRUE)
    
    np <- 200
    pred <- data.frame(visit_age = seq(min(df$visit_age, na.rm = TRUE),
                                       max(df$visit_age, na.rm = TRUE),
                                       length.out = np))
    
    ## Derivatives
    derv <- gratia::derivatives(gam.model, term = "s(visit_age)", 
                                data = pred, interval = "simultaneous", unconditional = FALSE)
    derv <- derv %>% mutate(sig = !(0 > .lower_ci & 0 < .upper_ci),
                            sig_deriv = .derivative * sig)
    
    mean.derivative <- mean(derv$.derivative)
    gam.smooth.F <- gam.results$s.table[1,3]
    gam.smooth.Rsq <- gam.results$r.sq
    if(mean.derivative < 0) gam.smooth.F <- -gam.smooth.F
    gam.smooth.pvalue <- gam.results$s.table[1,4]
    
    ## Temporal characteristics
    change.onset <- ifelse(any(derv$sig), min(derv$visit_age[derv$sig]), NA)
    peak.change <- ifelse(any(derv$sig),
                          mean(derv$visit_age[which.max(abs(derv$sig_deriv))]),
                          NA)
    decrease.onset <- ifelse(any(derv$sig_deriv < 0), min(derv$visit_age[derv$sig_deriv < 0]), NA)
    decrease.offset <- ifelse(any(derv$sig_deriv < 0), max(derv$visit_age[derv$sig_deriv < 0]), NA)
    increase.onset <- ifelse(any(derv$sig_deriv > 0), min(derv$visit_age[derv$sig_deriv > 0]), NA)
    increase.offset <- ifelse(any(derv$sig_deriv > 0), max(derv$visit_age[derv$sig_deriv > 0]), NA)
    change.offset <- ifelse(any(derv$sig), max(derv$visit_age[derv$sig]), NA)
    
    ## Fitted values & maxval2
    gam.fittedvalues_temp <- gratia::fitted_values(gam.model, data = pred)
    gam.fittedvalues_temp$.fitted2 <- plogis(gam.fittedvalues_temp$.fitted) * 30
    
    maxval2 <- max(gam.fittedvalues_temp$.fitted, na.rm = TRUE)
    max.age.fitted <- gam.fittedvalues_temp$visit_age[which.min(gam.fittedvalues_temp$.fitted)]
    
    ## Save stats
    gam.statistics <- bind_rows(gam.statistics, data.frame(
      oROI = subst,
      GAM.smooth_Fvalue = gam.smooth.F,
      GAM.smooth_RSq = gam.smooth.Rsq,
      GAM.smooth.pvalue = gam.smooth.pvalue,
      smooth.change.onset = change.onset,
      smooth.peak.change = peak.change,
      smooth.decrease.onset = decrease.onset,
      smooth.decrease.offset = decrease.offset,
      smooth.increase.onset = increase.onset,
      smooth.increase.offset = increase.offset,
      smooth.last.change = change.offset,
      max.age.fitted = max.age.fitted,
      max.val.fitted = maxval2
    ))
    
    ## Save fitted values
    gam.fittedvalues <- bind_rows(gam.fittedvalues, data.frame(
      oROI = subst, 
      visit_age = gam.fittedvalues_temp$visit_age,
      fitted = gam.fittedvalues_temp$.fitted,
      fitted2 = gam.fittedvalues_temp$.fitted2,
      se = gam.fittedvalues_temp$.se,
      lower_ci = gam.fittedvalues_temp$.lower_ci,
      upper_ci = gam.fittedvalues_temp$.upper_ci
    ))
    
    ## Save smooth estimates
    gam.smoothvalues_temp <- gratia::smooth_estimates(gam.model, data = pred, smooth = "s(visit_age)")
    gam.smoothestimates <- bind_rows(gam.smoothestimates, data.frame(
      oROI = subst, 
      visit_age = gam.smoothvalues_temp$visit_age,
      estimate = gam.smoothvalues_temp$.estimate,
      se = gam.smoothvalues_temp$.se
    ))
    
    ## Save derivatives
    gam.derivatives <- bind_rows(gam.derivatives, data.frame(
      oROI = subst,
      visit_age = derv$visit_age,
      derivative = derv$.derivative,
      lower = derv$.lower_ci,
      upper = derv$.upper_ci,
      significant = derv$sig,
      significant_deriv = derv$sig_deriv
    ))
    
    ## Save predictions
    gam.pred_temp <- ggpredict(gam.model, terms=c("visit_age"))
    gam.pred <- bind_rows(gam.pred, data.frame(
      oROI = subst, 
      visit_age = gam.pred_temp$x,
      predicted = gam.pred_temp$predicted,
      se = gam.pred_temp$std.error,
      lower = gam.pred_temp$conf.low,
      upper = gam.pred_temp$conf.high
    ))
  }

# Average derivative per group
deriv_rate <- gam.derivatives %>%
  group_by(oROI) %>%
  dplyr::summarise(deriv_rate = mean(derivative, na.rm = TRUE), .groups = "drop")

gam.statistics_all <- left_join(gam.statistics, deriv_rate, by = c("oROI"))
gam.statistics_all$Label <- "All"
gam.fittedvalues_all <- gam.fittedvalues
gam.fittedvalues_all$Label <- "All"
gam.pred_all <- gam.pred
gam.pred_all$Label <- "All"
gam.derivatives_all <- gam.derivatives
gam.derivatives_all$Label <- "All"

``` 

###Sex 
```{r, echo = FALSE, fig.align='center'}

roi <- levels(Iron$oROI)
sexes <- levels(Iron$sex)

# Initialize output containers
gam.statistics <- data.frame(
  oROI = factor(),
  Label = factor(),
  GAM.smooth_Fvalue = numeric(),
  GAM.smooth_RSq = numeric(),
  GAM.smooth.pvalue = numeric(),
  smooth.change.onset = numeric(),
  smooth.peak.change = numeric(),
  smooth.decrease.onset = numeric(),
  smooth.decrease.offset = numeric(),
  smooth.increase.onset = numeric(),
  smooth.increase.offset = numeric(),
  smooth.last.change = numeric(),
  max.age.fitted = numeric(),
  max.val.fitted = numeric()
)

gam.fittedvalues <- data.frame(
  oROI = factor(),
  Label = factor(),
  visit_age = numeric(),
  fitted = numeric(),
  fitted2 = numeric(),
  se = numeric(),
  lower_ci = numeric(),
  upper_ci = numeric()
)

gam.smoothestimates <- data.frame(
  oROI = factor(),
  Label = factor(),
  visit_age = numeric(),
  estimate = numeric(),
  se = numeric()
)

gam.derivatives <- data.frame(
  oROI = factor(),
  Label = factor(),
  visit_age = numeric(),
  derivative = numeric(),
  lower = numeric(),
  upper = numeric(),
  significant = numeric(),
  significant_deriv = numeric()
)

gam.pred <- data.frame(
  oROI = factor(),
  Label = factor(),
  visit_age = numeric(),
  predicted = numeric(),
  se = numeric(),
  lower = numeric(),
  upper = numeric()
)

# ----------------------------
# Loop over ROI & sexes
# ----------------------------
for (subst in roi) {
  for (sx in sexes) {
    
    sub_df <- Iron %>%
      filter(oROI == subst, sex == sx)
    
    # if (nrow(sub_df) < 10) next  # skip if too few rows
    
    gam.model <- mgcv::gamm(
      tat2.combat ~ s(visit_age, k = 4, fx = F, bs = "cs"),
      random = list(subject = ~1, subject = ~0 + visit_age),
      data = sub_df,
      method = "REML"
    )
    
    gam.results <- summary(gam.model$gam)
    df <- as.data.frame(unclass(gam.model$gam$model), stringsAsFactors = TRUE)
    
    np <- 200
    pred <- data.frame(visit_age = seq(min(df$visit_age, na.rm = TRUE),
                                       max(df$visit_age, na.rm = TRUE),
                                       length.out = np))
    
    ## Derivatives
    derv <- gratia::derivatives(gam.model, term = "s(visit_age)", 
                                data = pred, interval = "simultaneous", unconditional = FALSE)
    derv <- derv %>% mutate(sig = !(0 > .lower_ci & 0 < .upper_ci),
                            sig_deriv = .derivative * sig)
    
    mean.derivative <- mean(derv$.derivative)
    gam.smooth.F <- gam.results$s.table[1,3]
    gam.smooth.Rsq <- gam.results$r.sq
    if(mean.derivative < 0) gam.smooth.F <- -gam.smooth.F
    gam.smooth.pvalue <- gam.results$s.table[1,4]
    
    ## Temporal characteristics
    change.onset <- ifelse(any(derv$sig), min(derv$visit_age[derv$sig]), NA)
    peak.change <- ifelse(any(derv$sig),
                          mean(derv$visit_age[which.max(abs(derv$sig_deriv))]),
                          NA)
    decrease.onset <- ifelse(any(derv$sig_deriv < 0), min(derv$visit_age[derv$sig_deriv < 0]), NA)
    decrease.offset <- ifelse(any(derv$sig_deriv < 0), max(derv$visit_age[derv$sig_deriv < 0]), NA)
    increase.onset <- ifelse(any(derv$sig_deriv > 0), min(derv$visit_age[derv$sig_deriv > 0]), NA)
    increase.offset <- ifelse(any(derv$sig_deriv > 0), max(derv$visit_age[derv$sig_deriv > 0]), NA)
    change.offset <- ifelse(any(derv$sig), max(derv$visit_age[derv$sig]), NA)
    
    ## Fitted values & maxval2
    gam.fittedvalues_temp <- gratia::fitted_values(gam.model, data = pred)
    gam.fittedvalues_temp$.fitted2 <- plogis(gam.fittedvalues_temp$.fitted) * 30
    
    maxval2 <- max(gam.fittedvalues_temp$.fitted, na.rm = TRUE)
    max.age.fitted <- gam.fittedvalues_temp$visit_age[which.min(gam.fittedvalues_temp$.fitted)]
    
    ## Save stats
    gam.statistics <- bind_rows(gam.statistics, data.frame(
      oROI = subst,
      Label = sx,
      GAM.smooth_Fvalue = gam.smooth.F,
      GAM.smooth_RSq = gam.smooth.Rsq,
      GAM.smooth.pvalue = gam.smooth.pvalue,
      smooth.change.onset = change.onset,
      smooth.peak.change = peak.change,
      smooth.decrease.onset = decrease.onset,
      smooth.decrease.offset = decrease.offset,
      smooth.increase.onset = increase.onset,
      smooth.increase.offset = increase.offset,
      smooth.last.change = change.offset,
      max.age.fitted = max.age.fitted,
      max.val.fitted = maxval2
    ))
    
    ## Save fitted values
    gam.fittedvalues <- bind_rows(gam.fittedvalues, data.frame(
      oROI = subst, Label = sx,
      visit_age = gam.fittedvalues_temp$visit_age,
      fitted = gam.fittedvalues_temp$.fitted,
      fitted2 = gam.fittedvalues_temp$.fitted2,
      se = gam.fittedvalues_temp$.se,
      lower_ci = gam.fittedvalues_temp$.lower_ci,
      upper_ci = gam.fittedvalues_temp$.upper_ci
    ))
    
    ## Save smooth estimates
    gam.smoothvalues_temp <- gratia::smooth_estimates(gam.model, data = pred, smooth = "s(visit_age)")
    gam.smoothestimates <- bind_rows(gam.smoothestimates, data.frame(
      oROI = subst, Label = sx,
      visit_age = gam.smoothvalues_temp$visit_age,
      estimate = gam.smoothvalues_temp$.estimate,
      se = gam.smoothvalues_temp$.se
    ))
    
    ## Save derivatives
    gam.derivatives <- bind_rows(gam.derivatives, data.frame(
      oROI = subst, Label = sx,
      visit_age = derv$visit_age,
      derivative = derv$.derivative,
      lower = derv$.lower_ci,
      upper = derv$.upper_ci,
      significant = derv$sig,
      significant_deriv = derv$sig_deriv
    ))
    
    ## Save predictions
    gam.pred_temp <- ggpredict(gam.model, terms=c("visit_age"))
    gam.pred <- bind_rows(gam.pred, data.frame(
      oROI = subst, Label = sx,
      visit_age = gam.pred_temp$x,
      predicted = gam.pred_temp$predicted,
      se = gam.pred_temp$std.error,
      lower = gam.pred_temp$conf.low,
      upper = gam.pred_temp$conf.high
    ))
  }
}

# Average derivative per group
deriv_rate <- gam.derivatives %>%
  group_by(oROI, Label) %>%
  dplyr::summarise(deriv_rate = mean(derivative, na.rm = TRUE), .groups = "drop")

gam.statistics_sex <- left_join(gam.statistics, deriv_rate, by = c("oROI","Label"))
gam.fittedvalues_sex <- gam.fittedvalues
gam.pred_sex <- gam.pred
gam.derivatives_sex <- gam.derivatives

``` 

###Income 
```{r, echo = FALSE, fig.align='center'}

roi <- levels(Iron$oROI)
incomes <- levels(Iron$oIncome_recode)

# Initialize output containers
gam.statistics <- data.frame(
  oROI = factor(),
  Label = factor(),
  GAM.smooth_Fvalue = numeric(),
  GAM.smooth_RSq = numeric(),
  GAM.smooth.pvalue = numeric(),
  smooth.change.onset = numeric(),
  smooth.peak.change = numeric(),
  smooth.decrease.onset = numeric(),
  smooth.decrease.offset = numeric(),
  smooth.increase.onset = numeric(),
  smooth.increase.offset = numeric(),
  smooth.last.change = numeric(),
  max.age.fitted = numeric(),
  max.val.fitted = numeric()
)

gam.fittedvalues <- data.frame(
  oROI = factor(),
  Label = factor(),
  visit_age = numeric(),
  fitted = numeric(),
  fitted2 = numeric(),
  se = numeric(),
  lower_ci = numeric(),
  upper_ci = numeric()
)

gam.smoothestimates <- data.frame(
  oROI = factor(),
  Label = factor(),
  visit_age = numeric(),
  estimate = numeric(),
  se = numeric()
)

gam.derivatives <- data.frame(
  oROI = factor(),
  Label = factor(),
  visit_age = numeric(),
  derivative = numeric(),
  lower = numeric(),
  upper = numeric(),
  significant = numeric(),
  significant_deriv = numeric()
)

gam.pred <- data.frame(
  oROI = factor(),
  Label = factor(),
  visit_age = numeric(),
  predicted = numeric(),
  se = numeric(),
  lower = numeric(),
  upper = numeric()
)

# ----------------------------
# Loop over ROI & incomes
# ----------------------------
for (subst in roi) {
  for (sx in incomes) {
    
    sub_df <- Iron %>%
      filter(oROI == subst, oIncome_recode == sx)
    
    # if (nrow(sub_df) < 10) next  # skip if too few rows
    
    gam.model <- mgcv::gamm(
      tat2.combat ~ s(visit_age, k = 4, fx = F, bs = "cs"),
      random = list(subject = ~1, subject = ~0 + visit_age),
      data = sub_df,
      method = "REML"
    )
    
    gam.results <- summary(gam.model$gam)
    df <- as.data.frame(unclass(gam.model$gam$model), stringsAsFactors = TRUE)
    
    np <- 200
    pred <- data.frame(visit_age = seq(min(df$visit_age, na.rm = TRUE),
                                       max(df$visit_age, na.rm = TRUE),
                                       length.out = np))
    
    ## Derivatives
    derv <- gratia::derivatives(gam.model, term = "s(visit_age)", 
                                data = pred, interval = "simultaneous", unconditional = FALSE)
    derv <- derv %>% mutate(sig = !(0 > .lower_ci & 0 < .upper_ci),
                            sig_deriv = .derivative * sig)
    
    mean.derivative <- mean(derv$.derivative)
    gam.smooth.F <- gam.results$s.table[1,3]
    gam.smooth.Rsq <- gam.results$r.sq
    if(mean.derivative < 0) gam.smooth.F <- -gam.smooth.F
    gam.smooth.pvalue <- gam.results$s.table[1,4]
    
    ## Temporal characteristics
    change.onset <- ifelse(any(derv$sig), min(derv$visit_age[derv$sig]), NA)
    peak.change <- ifelse(any(derv$sig),
                          mean(derv$visit_age[which.max(abs(derv$sig_deriv))]),
                          NA)
    decrease.onset <- ifelse(any(derv$sig_deriv < 0), min(derv$visit_age[derv$sig_deriv < 0]), NA)
    decrease.offset <- ifelse(any(derv$sig_deriv < 0), max(derv$visit_age[derv$sig_deriv < 0]), NA)
    increase.onset <- ifelse(any(derv$sig_deriv > 0), min(derv$visit_age[derv$sig_deriv > 0]), NA)
    increase.offset <- ifelse(any(derv$sig_deriv > 0), max(derv$visit_age[derv$sig_deriv > 0]), NA)
    change.offset <- ifelse(any(derv$sig), max(derv$visit_age[derv$sig]), NA)
    
    ## Fitted values & maxval2
    gam.fittedvalues_temp <- gratia::fitted_values(gam.model, data = pred)
    gam.fittedvalues_temp$.fitted2 <- plogis(gam.fittedvalues_temp$.fitted) * 30
    
    maxval2 <- max(gam.fittedvalues_temp$.fitted, na.rm = TRUE)
    max.age.fitted <- gam.fittedvalues_temp$visit_age[which.min(gam.fittedvalues_temp$.fitted)]
    
    ## Save stats
    gam.statistics <- bind_rows(gam.statistics, data.frame(
      oROI = subst,
      Label = sx,
      GAM.smooth_Fvalue = gam.smooth.F,
      GAM.smooth_RSq = gam.smooth.Rsq,
      GAM.smooth.pvalue = gam.smooth.pvalue,
      smooth.change.onset = change.onset,
      smooth.peak.change = peak.change,
      smooth.decrease.onset = decrease.onset,
      smooth.decrease.offset = decrease.offset,
      smooth.increase.onset = increase.onset,
      smooth.increase.offset = increase.offset,
      smooth.last.change = change.offset,
      max.age.fitted = max.age.fitted,
      max.val.fitted = maxval2
    ))
    
    ## Save fitted values
    gam.fittedvalues <- bind_rows(gam.fittedvalues, data.frame(
      oROI = subst, Label = sx,
      visit_age = gam.fittedvalues_temp$visit_age,
      fitted = gam.fittedvalues_temp$.fitted,
      fitted2 = gam.fittedvalues_temp$.fitted2,
      se = gam.fittedvalues_temp$.se,
      lower_ci = gam.fittedvalues_temp$.lower_ci,
      upper_ci = gam.fittedvalues_temp$.upper_ci
    ))
    
    ## Save smooth estimates
    gam.smoothvalues_temp <- gratia::smooth_estimates(gam.model, data = pred, smooth = "s(visit_age)")
    gam.smoothestimates <- bind_rows(gam.smoothestimates, data.frame(
      oROI = subst, Label = sx,
      visit_age = gam.smoothvalues_temp$visit_age,
      estimate = gam.smoothvalues_temp$.estimate,
      se = gam.smoothvalues_temp$.se
    ))
    
    ## Save derivatives
    gam.derivatives <- bind_rows(gam.derivatives, data.frame(
      oROI = subst, Label = sx,
      visit_age = derv$visit_age,
      derivative = derv$.derivative,
      lower = derv$.lower_ci,
      upper = derv$.upper_ci,
      significant = derv$sig,
      significant_deriv = derv$sig_deriv
    ))
    
    ## Save predictions
    gam.pred_temp <- ggpredict(gam.model, terms=c("visit_age"))
    gam.pred <- bind_rows(gam.pred, data.frame(
      oROI = subst, Label = sx,
      visit_age = gam.pred_temp$x,
      predicted = gam.pred_temp$predicted,
      se = gam.pred_temp$std.error,
      lower = gam.pred_temp$conf.low,
      upper = gam.pred_temp$conf.high
    ))
  }
}

# Average derivative per group
deriv_rate <- gam.derivatives %>%
  group_by(oROI, Label) %>%
  dplyr::summarise(deriv_rate = mean(derivative, na.rm = TRUE), .groups = "drop")

gam.statistics_income <- left_join(gam.statistics, deriv_rate, by = c("oROI","Label"))
gam.fittedvalues_income <- gam.fittedvalues
gam.pred_income <- gam.pred
gam.derivatives_income <- gam.derivatives

``` 

###Race 
```{r, echo = FALSE, fig.align='center'}

roi <- levels(Iron$oROI)
races <- levels(Iron$oRaceeth_recode)

# Initialize output containers
gam.statistics <- data.frame(
  oROI = factor(),
  Label = factor(),
  GAM.smooth_Fvalue = numeric(),
  GAM.smooth_RSq = numeric(),
  GAM.smooth.pvalue = numeric(),
  smooth.change.onset = numeric(),
  smooth.peak.change = numeric(),
  smooth.decrease.onset = numeric(),
  smooth.decrease.offset = numeric(),
  smooth.increase.onset = numeric(),
  smooth.increase.offset = numeric(),
  smooth.last.change = numeric(),
  max.age.fitted = numeric(),
  max.val.fitted = numeric()
)

gam.fittedvalues <- data.frame(
  oROI = factor(),
  Label = factor(),
  visit_age = numeric(),
  fitted = numeric(),
  fitted2 = numeric(),
  se = numeric(),
  lower_ci = numeric(),
  upper_ci = numeric()
)

gam.smoothestimates <- data.frame(
  oROI = factor(),
  Label = factor(),
  visit_age = numeric(),
  estimate = numeric(),
  se = numeric()
)

gam.derivatives <- data.frame(
  oROI = factor(),
  Label = factor(),
  visit_age = numeric(),
  derivative = numeric(),
  lower = numeric(),
  upper = numeric(),
  significant = numeric(),
  significant_deriv = numeric()
)

gam.pred <- data.frame(
  oROI = factor(),
  Label = factor(),
  visit_age = numeric(),
  predicted = numeric(),
  se = numeric(),
  lower = numeric(),
  upper = numeric()
)

# ----------------------------
# Loop over ROI & incomes
# ----------------------------
for (subst in roi) {
  for (sx in races) {
    
    sub_df <- Iron %>%
      filter(oROI == subst, oRaceeth_recode == sx)
    
    # if (nrow(sub_df) < 10) next  # skip if too few rows
    
    gam.model <- mgcv::gamm(
      tat2.combat ~ s(visit_age, k = 4, fx = F, bs = "cs"),
      random = list(subject = ~1, subject = ~0 + visit_age),
      data = sub_df,
      method = "REML"
    )
    
    gam.results <- summary(gam.model$gam)
    df <- as.data.frame(unclass(gam.model$gam$model), stringsAsFactors = TRUE)
    
    np <- 200
    pred <- data.frame(visit_age = seq(min(df$visit_age, na.rm = TRUE),
                                       max(df$visit_age, na.rm = TRUE),
                                       length.out = np))
    
    ## Derivatives
    derv <- gratia::derivatives(gam.model, term = "s(visit_age)", 
                                data = pred, interval = "simultaneous", unconditional = FALSE)
    derv <- derv %>% mutate(sig = !(0 > .lower_ci & 0 < .upper_ci),
                            sig_deriv = .derivative * sig)
    
    mean.derivative <- mean(derv$.derivative)
    gam.smooth.F <- gam.results$s.table[1,3]
    gam.smooth.Rsq <- gam.results$r.sq
    if(mean.derivative < 0) gam.smooth.F <- -gam.smooth.F
    gam.smooth.pvalue <- gam.results$s.table[1,4]
    
    ## Temporal characteristics
    change.onset <- ifelse(any(derv$sig), min(derv$visit_age[derv$sig]), NA)
    peak.change <- ifelse(any(derv$sig),
                          mean(derv$visit_age[which.max(abs(derv$sig_deriv))]),
                          NA)
    decrease.onset <- ifelse(any(derv$sig_deriv < 0), min(derv$visit_age[derv$sig_deriv < 0]), NA)
    decrease.offset <- ifelse(any(derv$sig_deriv < 0), max(derv$visit_age[derv$sig_deriv < 0]), NA)
    increase.onset <- ifelse(any(derv$sig_deriv > 0), min(derv$visit_age[derv$sig_deriv > 0]), NA)
    increase.offset <- ifelse(any(derv$sig_deriv > 0), max(derv$visit_age[derv$sig_deriv > 0]), NA)
    change.offset <- ifelse(any(derv$sig), max(derv$visit_age[derv$sig]), NA)
    
    ## Fitted values & maxval2
    gam.fittedvalues_temp <- gratia::fitted_values(gam.model, data = pred)
    gam.fittedvalues_temp$.fitted2 <- plogis(gam.fittedvalues_temp$.fitted) * 30
    
    maxval2 <- max(gam.fittedvalues_temp$.fitted, na.rm = TRUE)
    max.age.fitted <- gam.fittedvalues_temp$visit_age[which.min(gam.fittedvalues_temp$.fitted)]
    
    ## Save stats
    gam.statistics <- bind_rows(gam.statistics, data.frame(
      oROI = subst,
      Label = sx,
      GAM.smooth_Fvalue = gam.smooth.F,
      GAM.smooth_RSq = gam.smooth.Rsq,
      GAM.smooth.pvalue = gam.smooth.pvalue,
      smooth.change.onset = change.onset,
      smooth.peak.change = peak.change,
      smooth.decrease.onset = decrease.onset,
      smooth.decrease.offset = decrease.offset,
      smooth.increase.onset = increase.onset,
      smooth.increase.offset = increase.offset,
      smooth.last.change = change.offset,
      max.age.fitted = max.age.fitted,
      max.val.fitted = maxval2
    ))
    
    ## Save fitted values
    gam.fittedvalues <- bind_rows(gam.fittedvalues, data.frame(
      oROI = subst, Label = sx,
      visit_age = gam.fittedvalues_temp$visit_age,
      fitted = gam.fittedvalues_temp$.fitted,
      fitted2 = gam.fittedvalues_temp$.fitted2,
      se = gam.fittedvalues_temp$.se,
      lower_ci = gam.fittedvalues_temp$.lower_ci,
      upper_ci = gam.fittedvalues_temp$.upper_ci
    ))
    
    ## Save smooth estimates
    gam.smoothvalues_temp <- gratia::smooth_estimates(gam.model, data = pred, smooth = "s(visit_age)")
    gam.smoothestimates <- bind_rows(gam.smoothestimates, data.frame(
      oROI = subst, Label = sx,
      visit_age = gam.smoothvalues_temp$visit_age,
      estimate = gam.smoothvalues_temp$.estimate,
      se = gam.smoothvalues_temp$.se
    ))
    
    ## Save derivatives
    gam.derivatives <- bind_rows(gam.derivatives, data.frame(
      oROI = subst, Label = sx,
      visit_age = derv$visit_age,
      derivative = derv$.derivative,
      lower = derv$.lower_ci,
      upper = derv$.upper_ci,
      significant = derv$sig,
      significant_deriv = derv$sig_deriv
    ))
    
    ## Save predictions
    gam.pred_temp <- ggpredict(gam.model, terms=c("visit_age"))
    gam.pred <- bind_rows(gam.pred, data.frame(
      oROI = subst, Label = sx,
      visit_age = gam.pred_temp$x,
      predicted = gam.pred_temp$predicted,
      se = gam.pred_temp$std.error,
      lower = gam.pred_temp$conf.low,
      upper = gam.pred_temp$conf.high
    ))
  }
}

# Average derivative per group
deriv_rate <- gam.derivatives %>%
  group_by(oROI, Label) %>%
  dplyr::summarise(deriv_rate = mean(derivative, na.rm = TRUE), .groups = "drop")

gam.statistics_raceeth <- left_join(gam.statistics, deriv_rate, by = c("oROI","Label"))
gam.fittedvalues_raceeth <- gam.fittedvalues
gam.pred_raceeth <- gam.pred
gam.derivatives_raceeth <- gam.derivatives

``` 


### Site
```{r, echo = FALSE, fig.align='center'}

roi <- levels(Iron$oROI)
sites <- levels(Iron$site)

# Initialize output containers
gam.statistics <- data.frame(
  oROI = factor(),
  Label = factor(),
  GAM.smooth_Fvalue = numeric(),
  GAM.smooth_RSq = numeric(),
  GAM.smooth.pvalue = numeric(),
  smooth.change.onset = numeric(),
  smooth.peak.change = numeric(),
  smooth.decrease.onset = numeric(),
  smooth.decrease.offset = numeric(),
  smooth.increase.onset = numeric(),
  smooth.increase.offset = numeric(),
  smooth.last.change = numeric(),
  max.age.fitted = numeric(),
  max.val.fitted = numeric()
)

gam.fittedvalues <- data.frame(
  oROI = factor(),
  Label = factor(),
  visit_age = numeric(),
  fitted = numeric(),
  fitted2 = numeric(),
  se = numeric(),
  lower_ci = numeric(),
  upper_ci = numeric()
)

gam.smoothestimates <- data.frame(
  oROI = factor(),
  Label = factor(),
  visit_age = numeric(),
  estimate = numeric(),
  se = numeric()
)

gam.derivatives <- data.frame(
  oROI = factor(),
  Label = factor(),
  visit_age = numeric(),
  derivative = numeric(),
  lower = numeric(),
  upper = numeric(),
  significant = numeric(),
  significant_deriv = numeric()
)

gam.pred <- data.frame(
  oROI = factor(),
  Label = factor(),
  visit_age = numeric(),
  predicted = numeric(),
  se = numeric(),
  lower = numeric(),
  upper = numeric()
)

# ----------------------------
# Loop over ROI & incomes
# ----------------------------
for (subst in roi) {
  for (sx in sites) {
    
    sub_df <- Iron %>%
      filter(oROI == subst, site == sx)
    
    # if (nrow(sub_df) < 10) next  # skip if too few rows
    
    gam.model <- mgcv::gamm(
      tat2.combat ~ s(visit_age, k = 4, fx = F, bs = "cs"),
      random = list(subject = ~1, subject = ~0 + visit_age),
      data = sub_df,
      method = "REML"
    )
    
    gam.results <- summary(gam.model$gam)
    df <- as.data.frame(unclass(gam.model$gam$model), stringsAsFactors = TRUE)
    
    np <- 200
    pred <- data.frame(visit_age = seq(min(df$visit_age, na.rm = TRUE),
                                       max(df$visit_age, na.rm = TRUE),
                                       length.out = np))
    
    ## Derivatives
    derv <- gratia::derivatives(gam.model, term = "s(visit_age)", 
                                data = pred, interval = "simultaneous", unconditional = FALSE)
    derv <- derv %>% mutate(sig = !(0 > .lower_ci & 0 < .upper_ci),
                            sig_deriv = .derivative * sig)
    
    mean.derivative <- mean(derv$.derivative)
    gam.smooth.F <- gam.results$s.table[1,3]
    gam.smooth.Rsq <- gam.results$r.sq
    if(mean.derivative < 0) gam.smooth.F <- -gam.smooth.F
    gam.smooth.pvalue <- gam.results$s.table[1,4]
    
    ## Temporal characteristics
    change.onset <- ifelse(any(derv$sig), min(derv$visit_age[derv$sig]), NA)
    peak.change <- ifelse(any(derv$sig),
                          mean(derv$visit_age[which.max(abs(derv$sig_deriv))]),
                          NA)
    decrease.onset <- ifelse(any(derv$sig_deriv < 0), min(derv$visit_age[derv$sig_deriv < 0]), NA)
    decrease.offset <- ifelse(any(derv$sig_deriv < 0), max(derv$visit_age[derv$sig_deriv < 0]), NA)
    increase.onset <- ifelse(any(derv$sig_deriv > 0), min(derv$visit_age[derv$sig_deriv > 0]), NA)
    increase.offset <- ifelse(any(derv$sig_deriv > 0), max(derv$visit_age[derv$sig_deriv > 0]), NA)
    change.offset <- ifelse(any(derv$sig), max(derv$visit_age[derv$sig]), NA)
    
    ## Fitted values & maxval2
    gam.fittedvalues_temp <- gratia::fitted_values(gam.model, data = pred)
    gam.fittedvalues_temp$.fitted2 <- plogis(gam.fittedvalues_temp$.fitted) * 30
    
    maxval2 <- max(gam.fittedvalues_temp$.fitted, na.rm = TRUE)
    max.age.fitted <- gam.fittedvalues_temp$visit_age[which.min(gam.fittedvalues_temp$.fitted)]
    
    ## Save stats
    gam.statistics <- bind_rows(gam.statistics, data.frame(
      oROI = subst,
      Label = sx,
      GAM.smooth_Fvalue = gam.smooth.F,
      GAM.smooth_RSq = gam.smooth.Rsq,
      GAM.smooth.pvalue = gam.smooth.pvalue,
      smooth.change.onset = change.onset,
      smooth.peak.change = peak.change,
      smooth.decrease.onset = decrease.onset,
      smooth.decrease.offset = decrease.offset,
      smooth.increase.onset = increase.onset,
      smooth.increase.offset = increase.offset,
      smooth.last.change = change.offset,
      max.age.fitted = max.age.fitted,
      max.val.fitted = maxval2
    ))
    
    ## Save fitted values
    gam.fittedvalues <- bind_rows(gam.fittedvalues, data.frame(
      oROI = subst, Label = sx,
      visit_age = gam.fittedvalues_temp$visit_age,
      fitted = gam.fittedvalues_temp$.fitted,
      fitted2 = gam.fittedvalues_temp$.fitted2,
      se = gam.fittedvalues_temp$.se,
      lower_ci = gam.fittedvalues_temp$.lower_ci,
      upper_ci = gam.fittedvalues_temp$.upper_ci
    ))
    
    ## Save smooth estimates
    gam.smoothvalues_temp <- gratia::smooth_estimates(gam.model, data = pred, smooth = "s(visit_age)")
    gam.smoothestimates <- bind_rows(gam.smoothestimates, data.frame(
      oROI = subst, Label = sx,
      visit_age = gam.smoothvalues_temp$visit_age,
      estimate = gam.smoothvalues_temp$.estimate,
      se = gam.smoothvalues_temp$.se
    ))
    
    ## Save derivatives
    gam.derivatives <- bind_rows(gam.derivatives, data.frame(
      oROI = subst, Label = sx,
      visit_age = derv$visit_age,
      derivative = derv$.derivative,
      lower = derv$.lower_ci,
      upper = derv$.upper_ci,
      significant = derv$sig,
      significant_deriv = derv$sig_deriv
    ))
    
    ## Save predictions
    gam.pred_temp <- ggpredict(gam.model, terms=c("visit_age"))
    gam.pred <- bind_rows(gam.pred, data.frame(
      oROI = subst, Label = sx,
      visit_age = gam.pred_temp$x,
      predicted = gam.pred_temp$predicted,
      se = gam.pred_temp$std.error,
      lower = gam.pred_temp$conf.low,
      upper = gam.pred_temp$conf.high
    ))
  }
}

# Average derivative per group
deriv_rate <- gam.derivatives %>%
  group_by(oROI, Label) %>%
  dplyr::summarise(deriv_rate = mean(derivative, na.rm = TRUE), .groups = "drop")

gam.statistics_site <- left_join(gam.statistics, deriv_rate, by = c("oROI","Label"))
gam.fittedvalues_site <- gam.fittedvalues
gam.pred_site <- gam.pred
gam.derivatives_site <- gam.derivatives

``` 

#### Combine the pred 
```{r, echo = FALSE, fig.align='center'}
 
All_Tat2_Fitted <- rbind(gam.fittedvalues_sex, gam.fittedvalues_income, gam.fittedvalues_raceeth, gam.fittedvalues_site, gam.fittedvalues_all)

All_Tat2_Pred <- rbind(gam.pred_sex, gam.pred_income, gam.pred_raceeth, gam.pred_site, gam.pred_all)

All_Tat2_Stats <- rbind(gam.statistics_sex, gam.statistics_income, gam.statistics_raceeth, gam.statistics_site, gam.statistics_all)



# All_Tat2_SES_Statistics[] <- lapply(All_Tat2_SES_Statistics, function(x) if (is.numeric(x)) round(x, 2) else x)

custom_order <- c("BG", "Caudate", "NAcc", "Pallidum", "Putamen")
# Convert to factor with levels in custom order
All_Tat2_Stats <- All_Tat2_Stats %>%
  mutate(oROI = factor(oROI, levels = custom_order)) %>%
  arrange(oROI)

write.csv(All_Tat2_Stats, file='/Users/ashparr/Library/Mobile Documents/com~apple~CloudDocs/Documents/Pitt/Manuscripts & Projects/NCANDA Project/DATA/Smooth_Statistics_Across_IronROIs_SES.csv', row.names = FALSE)

``` 

### Plots: 
``` {r, echo = FALSE, fig.align='center'}

All_Tat2_Stats$oLabel<-ordered(All_Tat2_Stats$Label,
levels=c("E", "D", "C","B", "A", "Other","Non-Hispanic Black/AA", "Non-Hispanic White", "Hispanic", 
         "Asian AIAN NHPI OTHER", "200k+","100k-199k","75k-99k","50k-74k", "25k-49k","<25k", 
         "M","F","All"))

All_Tat2_Pred$oLabel<-ordered(All_Tat2_Pred$Label,
levels=c("E", "D", "C","B", "A", "Other","Non-Hispanic Black/AA", "Non-Hispanic White", "Hispanic", 
         "Asian AIAN NHPI OTHER", "200k+","100k-199k","75k-99k","50k-74k", "25k-49k","<25k", 
         "M","F","All"))

col_scale_btc <- c("grey60","grey60","grey60","grey60","grey60","grey60",
                   "grey60","grey60","grey60","grey60","grey60","grey60",
                   "grey60","grey60","grey60","grey60","grey60","grey60",
                   "#461248")

# Age by Iron: All ROIs, no datapoints - across all SES factors, model predicted values 

age1_ses <- ggplot(All_Tat2_Pred,
                    aes(x=visit_age, y=predicted, color=oLabel))+
  geom_line(linewidth = 1) + facet_wrap(~oROI, scales="free")+
   scale_color_manual(values=col_scale_btc)+
   theme(aspect.ratio=.5, legend.position = "none",
                               strip.background = element_blank(),
         axis.text = element_text(size = 8, family = "Arial", color = c("black")),
  axis.title.x = element_text(size = 9, family ="Arial", color = c("black")),
  axis.title.y = element_text(size = 9, family ="Arial", color = c("black")),
  axis.line = element_line(linewidth = .2), 
  axis.ticks = element_line(linewidth = .2),
        panel.grid.major = element_blank(),panel.grid.minor=element_blank(),
        panel.background=element_blank()) +
  # scale_y_continuous(trans = "reverse")+ #Formatting best for all ROIs (different scales)
        scale_y_continuous(trans = "reverse",limits = c(.0145, .0125))+#Formatting best for basal ganglia (main)
scale_x_continuous(limits = c(12, 30), breaks = seq(15, 30, by = 5))+
 labs(x = "\nAge",
       y = "nT2*w\n")
print(age1_ses)

ggsave(age1_ses, filename = "/Users/ashparr/Documents/Pitt/Manuscripts & Projects/NCANDA Project/Results 2025/Iron_Age_AllROIs_SES.pdf", device = "pdf", dpi = 500,
       width = 8, height = 8, units = "in")

# Age of Peak Iron Identified by smooth fits in GAM models
age1_5_SES <- All_Tat2_Stats %>% 
  ggplot(aes(x = max.age.fitted, y = oLabel, fill = oLabel)) +
  geom_point(shape = 21, size = 5.5, color = "white") +
  facet_wrap(~oROI, scales = "free")+
    scale_color_manual(values=Demo_Colors)+  
   scale_fill_manual(values=Demo_Colors)+  
  labs(x = "\nAge of Peak nT2*w (Smooth)",
       y = "Subgroup\n")+ 
          # scale_x_continuous(limits = c(24, 30), breaks = seq(24, 30, by = 2))+ #Formatting best for basal ganglia (main)
        # scale_x_continuous(limits = c(24, 30), breaks = seq(24, 30, by = 2))+ #Formatting best for consistency across all ROI
  theme(aspect.ratio = 1.5,  
        strip.background = element_blank(),
  legend.position = "none",
  axis.text = element_text(size = 8, family = "Arial", color = c("black")),
  axis.title.x = element_text(size = 9, family ="Arial", color = c("black")),
  axis.title.y = element_text(size = 9, family ="Arial", color = c("black")),
  axis.line = element_line(linewidth = .2), 
  axis.ticks = element_line(linewidth = .2),
  panel.grid.major = element_blank(),
  panel.grid.minor=element_blank(),
  panel.background=element_blank())
print(age1_5_SES)

# Average derivative indicating rate of change (increase in this case)

age2_5_SES <- All_Tat2_Stats %>% 
  ggplot(aes(x = deriv_rate, y = oLabel, fill = oLabel)) +
  geom_point(shape = 21, size = 5.5, color = "white") +
  facet_wrap(~oROI, scales = "free")+
    scale_color_manual(values=Demo_Colors)+  
   scale_fill_manual(values=Demo_Colors)+  
  labs(x = "\nRate of Increase",
       y = "Subgroup\n")+
  scale_x_continuous(trans = "reverse", limits = c(0, -.00015))+ 
  theme(aspect.ratio = 1.5,
                strip.background = element_blank(),
  legend.position = "none",
  axis.text = element_text(size = 8, family = "Arial", color = c("black")),
  axis.title.x = element_text(size = 9, family ="Arial", color = c("black")),
  axis.title.y = element_text(size = 9, family ="Arial", color = c("black")),
  axis.line = element_line(linewidth = .2),
  axis.ticks = element_line(linewidth = .2),
  panel.grid.major = element_blank(),
  panel.grid.minor=element_blank(),
  panel.background=element_blank())
print(age2_5_SES)

# Age of maturation indicated by last age where derivative did not contain 0 
age3_5_SES <- All_Tat2_Stats %>% 
  ggplot(aes(x = smooth.last.change, y = oLabel, fill = oLabel)) +
  geom_point(shape = 21, size = 5.5, color = "white") +
  facet_wrap(~oROI, scales = "free")+
    scale_color_manual(values=Demo_Colors)+  
   scale_fill_manual(values=Demo_Colors)+  
xlab("\nAge of Maturation") +
  ylab("Subgroup\n") +
            scale_x_continuous(limits = c(18, 30), breaks = seq(20, 30, by = 5))+
  theme(aspect.ratio = 1.5,
                  strip.background = element_blank(),
  legend.position = "none",
  axis.text = element_text(size = 8, family = "Arial", color = c("black")),
  axis.title.x = element_text(size = 9, family ="Arial", color = c("black")),
  axis.title.y = element_text(size = 9, family ="Arial", color = c("black")),
  axis.line = element_line(linewidth = .2), 
  axis.ticks = element_line(linewidth = .2),
  panel.grid.major = element_blank(),
  panel.grid.minor=element_blank(),
  panel.background=element_blank())
print(age3_5_SES)


age4_5_SES <- All_Tat2_Stats %>% 
  ggplot(aes(x = smooth.peak.change, y = oLabel, fill = oLabel)) +
  geom_point(shape = 21, size = 5.5, color = "white") +
  facet_wrap(~oROI, scales = "free")+
    scale_color_manual(values=Demo_Colors)+  
   scale_fill_manual(values=Demo_Colors)+ 
  xlab("\nPeak Change Age") +
  ylab("Subgroup\n") +
scale_x_continuous(limits = c(12, 30), breaks = seq(15, 30, by = 5))+
  theme(aspect.ratio = 1.5,
                  strip.background = element_blank(),
  legend.position = "none",
  axis.text = element_text(size = 8, family = "Arial", color = c("black")),
  axis.title.x = element_text(size = 9, family ="Arial", color = c("black")),
  axis.title.y = element_text(size = 9, family ="Arial", color = c("black")),
  axis.line = element_line(linewidth = .2), 
  axis.ticks = element_line(linewidth = .2),
  panel.grid.major = element_blank(),
  panel.grid.minor=element_blank(),
  panel.background=element_blank())
print(age4_5_SES)


# Age at which significant increases (first age where derivative did not include 0) started 
age5_5_SES <- All_Tat2_Stats %>% 
  ggplot(aes(x = smooth.decrease.onset, y = oLabel, fill = oLabel)) +
  geom_point(shape = 21, size = 5.5, color = "white") +
  facet_wrap(~oROI, scales = "free")+
    scale_color_manual(values=Demo_Colors)+  
   scale_fill_manual(values=Demo_Colors)+  
  xlab("\nIncrease Onset (Age)") +
  ylab("Subgroup\n") +
              # scale_x_continuous(limits = c(12, 24), breaks = seq(12, 24, by = 4))+#Formatting best for consistency across ROIs
              scale_x_continuous(limits = c(12, 20), breaks = seq(12, 20, by = 4))+#Formatting best for basal ganglia (main)
  theme(aspect.ratio = 1.5,
                  strip.background = element_blank(),
  legend.position = "none",
  axis.text = element_text(size = 8, family = "Arial", color = c("black")),
  axis.title.x = element_text(size = 9, family ="Arial", color = c("black")),
  axis.title.y = element_text(size = 9, family ="Arial", color = c("black")),
  axis.line = element_line(linewidth = .2), 
  axis.ticks = element_line(linewidth = .2),
  panel.grid.major = element_blank(),
  panel.grid.minor=element_blank(),
  panel.background=element_blank())
print(age5_5_SES)


# Age at which significant increases (upper age where derivative included 0) ended 
age6_5_SES <- All_Tat2_Stats %>% 
  ggplot(aes(x = smooth.decrease.offset, y = oLabel, fill = oLabel)) +
  geom_point(shape = 21, size = 5.5, color = "white") +
facet_wrap(~oROI, scales = "free")+
    scale_color_manual(values=Demo_Colors)+  
   scale_fill_manual(values=Demo_Colors)+  
   labs(x = "\nIncrease Offset (Age)",
       y = "Subgroup\n")+
              scale_x_continuous(limits = c(24, 30), breaks = seq(24, 30, by = 2))+#Formatting best for basal ganglia (main)
                # scale_x_continuous(limits = c(18, 30), breaks = seq(20, 30, by = 5))+#Formatting best for consistency across ROIs
theme(aspect.ratio = 1.5,
  legend.position = "none",
            strip.background = element_blank(),
  axis.text = element_text(size = 8, family = "Arial", color = c("black")),
  axis.title.x = element_text(size = 9, family ="Arial", color = c("black")),
  axis.title.y = element_text(size = 9, family ="Arial", color = c("black")),
  axis.line = element_line(linewidth = .2), 
  axis.ticks = element_line(linewidth = .2),
  panel.grid.major = element_blank(),
  panel.grid.minor=element_blank(),
  panel.background=element_blank())
print(age6_5_SES)
 

ggsave(age1_5_SES, filename = "/Users/ashparr/Documents/Pitt/Manuscripts & Projects/NCANDA Project/Results 2025/Iron_Age_AllROIs_Gam_Stats_PeakPerf_SES.pdf", device = "pdf", dpi = 500,
       width = 8, height = 8, units = "in")#

ggsave(age2_5_SES, filename = "/Users/ashparr/Documents/Pitt/Manuscripts & Projects/NCANDA Project/Results 2025/Iron_Age_AllROIs_Gam_Stats_RateofInc_SES.pdf", device = "pdf", dpi = 500,
       width = 8, height = 8, units = "in")#

ggsave(age3_5_SES, filename = "/Users/ashparr/Documents/Pitt/Manuscripts & Projects/NCANDA Project/Results 2025/Iron_Age_AllROIs_Gam_Stats_MatureAge_SES.pdf", device = "pdf", dpi = 500,
       width = 8, height = 8, units = "in")#

ggsave(age4_5_SES, filename = "/Users/ashparr/Documents/Pitt/Manuscripts & Projects/NCANDA Project/Results 2025/Iron_Age_AllROIs_Gam_Stats_PeakChangeAge_SES.pdf", device = "pdf", dpi = 500,
       width = 8, height = 8, units = "in")#

ggsave(age5_5_SES, filename = "/Users/ashparr/Documents/Pitt/Manuscripts & Projects/NCANDA Project/Results 2025/Iron_Age_AllROIs_Gam_Stats_IncOnsetAge_SES.pdf", device = "pdf", dpi = 500,
       width = 8, height = 8, units = "in")#

ggsave(age6_5_SES, filename = "/Users/ashparr/Documents/Pitt/Manuscripts & Projects/NCANDA Project/Results 2025/Iron_Age_AllROIs_Gam_Stats_IncOffsetAge_SES.pdf", device = "pdf", dpi = 500,
       width = 8, height = 8, units = "in")#

``` 

# Substance Use Analyses
## Subset DFs
```{r, echo = FALSE, fig.align='center'}

set.seed(1) #for consistency  

SU_Vars_Iron <- All_Sites_Iron %>% 
  dplyr::select(subject, visit, sex, visit_age, site, income_recode,raceeth_recode,
                contains("past_30"),tat2.combat, roi, oROI)  

SU_Vars_Iron$roi <- as.factor(SU_Vars_Iron$roi)

### Subset DF 
SU_Vars_Iron_Long <- All_Sites_Iron_Long %>% 
  dplyr::select(subject, visit, sex, visit_age, site, income_recode,raceeth_recode,
                contains("past_30"),contains("log_past30"), tat2.combat, Substance, oSubstance, Category, Usetype, 
roi, oROI)  

SU_Vars_Iron_Long$roi <- as.factor(SU_Vars_Iron_Long$roi)
SU_Vars_Iron_Long$Substance <- as.factor(SU_Vars_Iron_Long$Substance)

``` 

### Continuous Variable Analysis: Substance use by Iron 
#### Test whether effect of UPPS (overall) varies by substance
```{r, echo = FALSE, fig.align='center'}

# see if effect of iron differs across substances 
GMM4_int <- bam(log_past30 ~ 1 + sex  +  income_recode + raceeth_recode + site + 
    s(visit_age, k = 4, fx = F, bs = "cs") + 
      oSubstance + 
      s(tat2.combat, k = 3, fx = F)+
    s(tat2.combat, oSubstance, bs = "sz", k =3, fx = F) +      
      s(subject, bs = "re") + s(subject, visit_age, bs = "re"), #implement random effects, 
    data = SU_Vars_Iron_Long %>% filter(oSubstance!="All Prob" & oROI=="BG"),
    discrete = TRUE,
    method = "fREML")
summary(GMM4_int)
anova(GMM4_int)

 
int_gam.model_plot <- ggpredict(GMM4_int, terms = c("tat2.combat","oSubstance"))
plot(int_gam.model_plot)


mm <- hypothesis_test(GMM4_int, 
                terms = c("tat2.combat","oSubstance"), test = NULL)

  

``` 


#### Substance use: Test for Associations with nT2*w Across all Regions and Substances  

```{r, echo = FALSE, fig.align='center'}

#Loop over trial types & substances to get gamm model statistics

roi <- levels(SU_Vars_Iron_Long$oROI)
substance_levels <- levels(SU_Vars_Iron_Long$oSubstance)

# Initialize output data frames
gam.statistics <- data.frame(
  oROI = factor(),
  Substance = factor(),
  GAM.smooth_Fvalue = numeric(),
  GAM.smooth_Rsq = numeric(),
  GAM.smooth.pvalue = numeric()
)

gam.fittedvalues <- data.frame(
  oROI = factor(), 
  Substance = factor(),
  corrat = numeric(), 
  fitted = numeric(), 
  fitted2 = numeric(), 
  se = numeric(), 
  lower_ci = numeric(), 
  upper_ci = numeric()
)

gam.smoothestimates <- data.frame(
  oROI = factor(), 
  Substance = factor(),
  corrat = numeric(), 
  estimate = numeric(), 
  se = numeric()
)

gam.pred <- data.frame(
  oROI = factor(), 
  Substance = factor(),
  corrat = numeric(), 
  predicted = numeric(), 
  se = numeric(),
  lower = numeric(), 
  upper = numeric()
)

# Nested loop: for each ROI, loop over each Substance
for (cog in roi) {
  for (subst in substance_levels) {
    
    sub_df <- SU_Vars_Iron_Long %>%
      filter(roi == cog, oSubstance == subst)
    
    # if (nrow(sub_df) == 0) next  # skip if no rows
    
    gam.model <- mgcv::gamm(
      log_past30 ~ s(tat2.combat, fx = F, k = 3) + 
        sex + income_recode + raceeth_recode + site + 
        s(visit_age, fx = F, k = 4, bs = "cs"), 
      random = list(subject = ~1, subject = ~0 + visit_age),
      data = sub_df,
      method = "REML"
    )
    
    gam.results <- summary(gam.model$gam)
    
    df <- as.data.frame(unclass(gam.model$gam$model), stringsAsFactors = TRUE)
    np <- 200
    thisPred <- data.frame(init = rep(0, np))
    theseVars <- attr(gam.model$gam$terms, "term.labels")
    varClasses <- attr(gam.model$gam$terms,"dataClasses")
    thisResp <- as.character(gam.model$gam$terms[[2]])
    smooth_var <- "tat2.combat"
    
    for (v in seq_along(theseVars)) {
      thisVar <- theseVars[[v]]
      thisClass <- varClasses[thisVar]
      if (thisVar == smooth_var) {
        thisPred[,smooth_var] <- seq(min(df[,smooth_var],na.rm=T),
                                     max(df[,smooth_var],na.rm=T),
                                     length.out = np)
      } else {
        switch(thisClass,
               "numeric" = {thisPred[,thisVar] = median(df[,thisVar], na.rm=TRUE)},
               "character" = {thisPred[,thisVar] = levels(df[,thisVar])[[1]]},
               "factor" = {thisPred[,thisVar] = levels(df[,thisVar])[[1]]},
               "ordered" = {thisPred[,thisVar] = levels(df[,thisVar])[[1]]}
        )
      }
    }
    pred <- thisPred %>% dplyr::select(-init)
    
    ## MODEL STATISTICS ##  
    gam.smooth.F <- gam.results$s.table[1,3]
    gam.smooth.Rsq <- gam.results$r.sq
    gam.smooth.pvalue <- gam.results$s.table[1,4]
    
    gam.statistics <- bind_rows(gam.statistics, data.frame(
      oROI = factor(cog, levels = roi),
      Substance = factor(subst, levels = substance_levels),
      GAM.smooth_Fvalue = gam.smooth.F,
      GAM.smooth_Rsq = gam.smooth.Rsq,
      GAM.smooth.pvalue = gam.smooth.pvalue
    ))
    
    ## FITTED VALUES ##
    gam.fittedvalues_temp <- gratia::fitted_values(object = gam.model, data = pred)
    gam.fittedvalues_temp$.fitted2 <- plogis(gam.fittedvalues_temp$.fitted) * 30
    
    temp_fit <- data.frame(
      oROI = factor(cog, levels = roi),
      oSubstance = factor(subst, levels = substance_levels),
      tat2.combat = gam.fittedvalues_temp$tat2.combat, 
      fitted = gam.fittedvalues_temp$.fitted, 
      fitted2 = gam.fittedvalues_temp$.fitted2, 
      se = gam.fittedvalues_temp$.se, 
      lower_ci = gam.fittedvalues_temp$.lower_ci, 
      upper_ci = gam.fittedvalues_temp$.upper_ci
    )
    gam.fittedvalues <- bind_rows(gam.fittedvalues, temp_fit)
    
    ## SMOOTH ESTIMATES ##
    gam.smoothvalues_temp <- gratia::smooth_estimates(object = gam.model, data = pred, smooth = sprintf('s(%s)', smooth_var))
    temp_smooth <- data.frame(
      oROI = factor(cog, levels = roi),
      oSubstance = factor(subst, levels = substance_levels),
      tat2.combat = gam.smoothvalues_temp$tat2.combat, 
      estimate = gam.smoothvalues_temp$.estimate,  
      se = gam.smoothvalues_temp$.se
    )
    gam.smoothestimates <- bind_rows(gam.smoothestimates, temp_smooth)
    
    ## GGPREDICT ##
    gam.pred_temp <- ggpredict(gam.model, terms = c("tat2.combat"))
    temp_pred <- data.frame(
      oROI = factor(cog, levels = roi),
      Substance = factor(subst, levels = substance_levels),
      tat2.combat = gam.pred_temp$x,
      predicted = gam.pred_temp$predicted, 
      se = gam.pred_temp$std.error,
      lower = gam.pred_temp$conf.low, 
      upper = gam.pred_temp$conf.high
    )
    gam.pred <- bind_rows(gam.pred, temp_pred)
  }
}


custom_order <- c("All Prob", "Alcohol", "Binge", "Cannabis", 
                  "Nicotine")

custom_order2 <- c("BG", "Caudate", "NAcc", "Pallidum", "Putamen")


gam.statistics <- gam.statistics %>%
  mutate(oROI = factor(oROI, levels = custom_order2),
         Substance = factor(Substance, levels = custom_order)) %>%
  arrange(Substance)


write.csv(gam.statistics, file='/Users/ashparr/Library/Mobile Documents/com~apple~CloudDocs/Documents/Pitt/Manuscripts & Projects/NCANDA Project/DATA/SU_Iron_Statistics_Across_Regions_FullSample.csv.csv', row.names = FALSE)

``` 

#### Plots:
```{r, echo = FALSE, fig.align='center'}

                                                              
gam.pred$Substance<-factor(gam.pred$Substance,
levels=c("All Prob","Alcohol","Binge","Cannabis","Nicotine"))

# nT2*w by Substance Use: All ROIs, all substances, no datapoints - model predicted values adjusting for age and relevant demographic factors 

SU_Iron <- ggplot(gam.pred,
                    aes(x=tat2.combat, y=predicted, color=Substance))+
  facet_wrap(~oROI,scales="free")+
    geom_smooth(method="gam", formula = y ~ s(x, k = 3, fx = FALSE), fullrange=F)+
   scale_color_manual(values=substance_colorbar)+
   scale_fill_manual(values=substance_colorbar)+
     theme(aspect.ratio=.5, legend.position = "none",
                                        strip.background = element_blank(),
         axis.text = element_text(size = 8, family = "Arial", color = c("black")),
  axis.title.x = element_text(size = 9, family ="Arial", color = c("black")),
  axis.title.y = element_text(size = 9, family ="Arial", color = c("black")),
  axis.line = element_line(linewidth = .2), 
  axis.ticks = element_line(linewidth = .2),
        panel.grid.major = element_blank(),panel.grid.minor=element_blank(),
        panel.background=element_blank()) +
   scale_x_continuous(trans = "reverse")+
  labs(x = "\nnT2*w Score",
       y = "Substance Use (log)\n")

print(SU_Iron)


ggsave(SU_Iron, filename = "/Users/ashparr/Documents/Pitt/Manuscripts & Projects/NCANDA Project/Results 2025/Iron_SU_Cont_AllROI.pdf", device = "pdf", dpi = 500, width = 8, height = 8, units = "in")#


gam.statistics$Substance <- ordered(gam.statistics$Substance, c("All Prob",
                                                                                        "Alcohol",
                                                                                        "Cannabis",
                                                                                        "Nicotine",
                                                                                        "Binge"))
# Plots showing gam model F statistics for each comparison (substance)
F1_5 <- gam.statistics %>% 
  ggplot(aes(x = GAM.smooth_Fvalue, y = Substance, fill = Substance)) +
  facet_wrap(~oROI, scales = "free")+
  geom_point(shape = 21, size = 5.5, color = "white") +
  scale_fill_manual(values = substance_colorbar2) +
  labs(x = "\nF Statistic",
       y = "Substance\n") +
   scale_y_discrete(
    breaks = c("All Prob", "Alcohol", "Cannabis", "Nicotine", "Binge"),
    labels = c("All", "Alcohol", "Cannabis", "Nicotine", "Binge"))+ 
  scale_x_continuous(breaks = c(0, 5, 10), limits = c(0, 11)) +
  theme(aspect.ratio = 1.5,
                        strip.background = element_blank(),
  legend.position = "none",
  axis.text = element_text(size = 8, family = "Arial", color = c("black")),
  axis.title.x = element_text(size = 9, family ="Arial", color = c("black")),
  axis.title.y = element_text(size = 9, family ="Arial", color = c("black")),
  axis.line = element_line(linewidth = .2), 
  axis.ticks = element_line(linewidth = .2),
  panel.grid.major = element_blank(),
  panel.grid.minor=element_blank(),
  panel.background=element_blank())
print(F1_5)

ggsave(F1_5, filename = "/Users/ashparr/Documents/Pitt/Manuscripts & Projects/NCANDA Project/Results 2025/Iron_SU_AllROIs_Fval.pdf", device = "pdf", dpi = 500,
       width = 8, height = 8, units = "in")#


``` 

### Frequency Category Analysis: Substance use by Iron 
```{r, echo = FALSE, fig.align='center'}


SU_Vars_Cat_Long <- SU_Vars_Iron_Long %>% dplyr::select(subject, visit, sex, visit_age, site, income_recode,raceeth_recode,tat2.combat, oROI, oSubstance, Category)

SU_Results.Iron <- data.frame(
  oROI = factor(),
  oSubstance = factor(),
  F_value = numeric(),
  p_value = numeric(),
  stringsAsFactors = FALSE
)

gam.pred.Iron <- data.frame(
  oROI = factor(),
  oSubstance = factor(),
  Drug_Category = factor(),
  predicted = numeric(),
  se = numeric(),
  lower = numeric(),
  upper = numeric()
)

# Loop through ROI  Substance
for (s in unique(SU_Vars_Cat_Long$oROI)) {
  for (sub in unique(SU_Vars_Cat_Long$oSubstance)) {
    
    # Subset data for this combo
    sub_df <- SU_Vars_Cat_Long %>% 
      filter(oROI == s, oSubstance == sub)
    
    # if (nrow(sub_df) == 0) next  # skip if no rows
    
    # Fit GAMM
    model <- mgcv::gamm(
      tat2.combat ~ s(visit_age, k = 4, fx = FALSE, bs = "cs") + Category + income_recode + 
        raceeth_recode + sex + site,
      random = list(subject = ~1, subject = ~0 + visit_age),
      data = sub_df,
      method = "REML"
    )
    
    # Extract ANOVA results
    aov_result <- anova(model$gam)
    F_val <- aov_result$pTerms.table[1, 2]
    p_val <- aov_result$pTerms.table[1, 3]
    
    SU_Results.Iron <- rbind(
      SU_Results.Iron,
      data.frame(
        oROI = s,
        oSubstance = sub,
        F_value = round(F_val, 2),
        p_value = signif(p_val, 2),
        stringsAsFactors = FALSE
      )
    )
    
    # Predictions for "Category"
    get_preds <- ggpredict(model$gam, terms = c("Category"))
    
    gam.pred.Iron <- rbind(
      gam.pred.Iron,
      data.frame(
        oROI = s,
        oSubstance = sub,
        Drug_Category = get_preds$x,
        predicted = get_preds$predicted,
        se = get_preds$std.error,
        lower = get_preds$conf.low,
        upper = get_preds$conf.high
      )
    )
  }
}

custom_order <- c("BG", "Caudate", "NAcc", "Pallidum", "Putamen")

custom_order2 <- c("Never", "Ever", "Regular")

custom_order3 <- c("All Prob", "Alcohol", "Binge","Cannabis","Nicotine")

# Convert to factor with levels in custom order
SU_Results.Iron_All <- SU_Results.Iron %>%
  mutate(oROI = factor(oROI, levels = custom_order)) %>%
  mutate(oSubstance = factor(oSubstance, levels = custom_order3)) %>%
  arrange(oSubstance, oROI)

gam.pred.Iron_All <- gam.pred.Iron %>%
  mutate(Drug_Category = factor(Drug_Category, levels = custom_order2),
         oROI = factor(oROI, levels = custom_order)) %>%
    mutate(oSubstance = factor(oSubstance, levels = custom_order3)) %>%
  arrange(oSubstance, oROI, Drug_Category)


write.csv(SU_Results.Iron_All, file='/Users/ashparr/Library/Mobile Documents/com~apple~CloudDocs/Documents/Pitt/Manuscripts & Projects/NCANDA Project/DATA/Mean_Main_Effects_FreqCategories_Across_IronROIs.csv', row.names = FALSE)

write.csv(gam.pred.Iron_All, file='/Users/ashparr/Library/Mobile Documents/com~apple~CloudDocs/Documents/Pitt/Manuscripts & Projects/NCANDA Project/DATA/Mean_Predicted_FreqCategories_Across_IronROIs.csv', row.names = FALSE)



``` 

#### Plots: 
```{r, echo = FALSE, fig.align='center'}
  


# Plots showing nT2*w by substance use category groups - model predictions adjusted for age and relevant covariates

Freq1_5 <- gam.pred.Iron_All %>% 
  ggplot(aes(x = Drug_Category, y = predicted, fill = oSubstance)) +
  facet_wrap(~oROI, scales = "free")+
  geom_point(shape = 21, size = 2.5, color = "white",position = position_dodge(0.3)) +
  scale_fill_manual(values = substance_colorbar) +
  labs(x = "\nFrequency",
       y = "nT2*w\n")+
  # scale_y_continuous(trans = "reverse")+ #Formatting best for all ROIs (different scales)
      scale_y_continuous(trans = "reverse",limits = c(.01393, .01378))+#Formatting best for basal ganglia (main)
  theme(aspect.ratio = .5,
                strip.background = element_blank(),
  legend.position = "none",
  axis.text = element_text(size = 8, family = "Arial", color = c("black")),
  axis.title.x = element_text(size = 9, family ="Arial", color = c("black")),
  axis.title.y = element_text(size = 9, family ="Arial", color = c("black")),
  axis.line = element_line(linewidth = .2), 
  axis.ticks = element_line(linewidth = .2),
  panel.grid.major = element_blank(),
  panel.grid.minor=element_blank(),
  panel.background=element_blank())
print(Freq1_5)

ggsave(Freq1_5, filename = "/Users/ashparr/Documents/Pitt/Manuscripts & Projects/NCANDA Project/Results 2025/Iron_SU_FreqCat_AllROIs.pdf", device = "pdf", dpi = 500,
       width = 8, height = 8, units = "in")#
   

``` 

### Usetype Category Analysis: Substance use by UPPS 
```{r, echo = FALSE, fig.align='center'}
  

SU_Vars_Use_Long <- SU_Vars_Iron_Long %>% dplyr::select(subject, visit, sex, visit_age, site, income_recode,raceeth_recode,tat2.combat, oROI, oSubstance, Usetype)


SU_UT_Results.Iron <- data.frame(
  oROI = factor(),
  oSubstance = factor(),
  F_value = numeric(),
  p_value = numeric(),
  stringsAsFactors = FALSE
)

gam.pred.UT_Iron <- data.frame(
  oROI = factor(),
  oSubstance = factor(),
  Drug_Category = factor(),
  predicted = numeric(),
  se = numeric(),
  lower = numeric(),
  upper = numeric()
)

# Loop through ROI  Substance
for (s in unique(SU_Vars_Use_Long$oROI)) {
  for (sub in unique(SU_Vars_Use_Long$oSubstance)) {
    
    # Subset data for this combo
    sub_df <- SU_Vars_Use_Long %>% 
      filter(oROI == s, oSubstance == sub)
    
    # if (nrow(sub_df) == 0) next  # skip if no rows
    
    # Fit GAMM
    model <- mgcv::gamm(
      tat2.combat ~ s(visit_age, k = 4, fx = FALSE, bs = "cs") + Usetype + income_recode + 
        raceeth_recode + sex + site,
      random = list(subject = ~1, subject = ~0 + visit_age),
      data = sub_df,
      method = "REML"
    )
    
    # Extract ANOVA results
    aov_result <- anova(model$gam)
    F_val <- aov_result$pTerms.table[1, 2]
    p_val <- aov_result$pTerms.table[1, 3]
    
    SU_UT_Results.Iron <- rbind(
      SU_UT_Results.Iron,
      data.frame(
        oROI = s,
        oSubstance = sub,
        F_value = round(F_val, 2),
        p_value = signif(p_val, 2),
        stringsAsFactors = FALSE
      )
    )    

    
    # Predictions for "Usetype"
    get_preds <- ggpredict(model$gam, terms = c("Usetype"))
    
    gam.pred.UT_Iron <- rbind(
      gam.pred.UT_Iron,
      data.frame(
        oROI = s,
        oSubstance = sub,
        Drug_Category = get_preds$x,
        predicted = get_preds$predicted,
        se = get_preds$std.error,
        lower = get_preds$conf.low,
        upper = get_preds$conf.high
      )
    )
  }
}

custom_order <- c("BG", "Caudate", "NAcc", "Pallidum", "Putamen")

custom_order2 <- c("no-use", "single-use", "co-use", "poly-use")

custom_order3 <- c("All Prob", "Alcohol", "Binge","Cannabis","Nicotine")

# Convert to factor with levels in custom order
SU_UT_Results.Iron_All <- SU_UT_Results.Iron %>%
  mutate(oROI = factor(oROI, levels = custom_order)) %>%
  mutate(oSubstance = factor(oSubstance, levels = custom_order3)) %>%
  arrange(oSubstance, oROI)

gam.pred.UT_Iron_All <- gam.pred.UT_Iron %>%
  mutate(Drug_Category = factor(Drug_Category, levels = custom_order2),
         oROI = factor(oROI, levels = custom_order)) %>%
    mutate(oSubstance = factor(oSubstance, levels = custom_order3)) %>%
  arrange(oSubstance, oROI, Drug_Category)


write.csv(SU_UT_Results.Iron_All, file='/Users/ashparr/Library/Mobile Documents/com~apple~CloudDocs/Documents/Pitt/Manuscripts & Projects/NCANDA Project/DATA/Mean_Main_Effects_UsetypeCategories_Across_IronROIs.csv', row.names = FALSE)

write.csv(gam.pred.UT_Iron_All, file='/Users/ashparr/Library/Mobile Documents/com~apple~CloudDocs/Documents/Pitt/Manuscripts & Projects/NCANDA Project/DATA/Mean_Predicted_UsetypeCategories_Across_IronROIs.csv', row.names = FALSE)

``` 



#### Plots: 
```{r, echo = FALSE, fig.align='center'}
  

# Plots showing Iron by substance use use-type groups - model predictions adjusted for age and relevant covariates


Usetype1_5 <- gam.pred.UT_Iron_All %>% 
  ggplot(aes(x = Drug_Category, y = predicted, fill = oSubstance)) +
  facet_wrap(~oROI, scales = "free")+
  geom_point(shape = 21, size = 2.5, color = "white",position = position_dodge(0.3)) +
  scale_fill_manual(values = substance_colorbar) +
  labs(x = "\nUse-Type",
       y = "nT2*w\n")+
      # scale_y_continuous(trans = "reverse"), #Formatting best for consistency among ROIs.
        scale_y_continuous(trans = "reverse",limits = c(.01395, .01376))+# #Formatting best for basal ganglia (main)
  theme(aspect.ratio = .5,
  legend.position = "none",
                  strip.background = element_blank(),
  axis.text = element_text(size = 8, family = "Arial", color = c("black")),
  axis.title.x = element_text(size = 9, family ="Arial", color = c("black")),
  axis.title.y = element_text(size = 9, family ="Arial", color = c("black")),
  axis.line = element_line(linewidth = .2), 
  axis.ticks = element_line(linewidth = .2),
  panel.grid.major = element_blank(),
  panel.grid.minor=element_blank(),
  panel.background=element_blank())
print(Usetype1_5)

ggsave(Usetype1_5, filename = "/Users/ashparr/Documents/Pitt/Manuscripts & Projects/NCANDA Project/Results 2025/Iron_SU_UseTypeCat_AllROIs.pdf", device = "pdf", dpi = 500,
       width = 8, height = 8, units = "in")#
   

``` 
